---
title: "3.4. Lazy Evaluation and Lexical Scoping"
output:
     slidy_presentation
---

How is a function call actually executed? From the language
definition:

> The evaluator finds the function object being called, creates an
> environment with the formal arguments of the function matched to
> the actual arguments in the call, and evaluates the body of the
> function in that environment.

Lazy evaluation refers to the fact that _function arguments are not
evaluated until needed_.

> The objects in the environment of the call are not the evaluated
> objects, but _promises_.  They contain the expression, it's
> value (when done) and some flags.  The argument is evaluated when
> it is  needed. The promise remains a promise after evaluation but
> the evaluation is done only once.

Lazy evaluation is implemented using _promises_.

No way to see promises in R directly, but they enable things like
`substitute()` for parsing and checking for missing arguments.
(Note. There is an R package providing promises for asynchronous
programming, and although somewhat related, that is not exactly the
same.)

## An exercise

Consider:

```{r}
f <- function(x, y = x) {
    y
}
g <- function(x, y = x) {
  x <- 2 * x
  y
}
```

What will be returned for `f(10)` and `g(10)`?

```{r, eval = FALSE}
f(10)
g(10)
```

__Answer__: For `f(10)`, you get the expected `10` since when `y` is
evaluated, the value of `x` is looked up and assigned in the promise
associated with `y`.

For `g(10)`, the returned value is `20` because when `y` is lazily
evaluated, the value of `x`, has changed to `20`! Recall thet `y` is
implemented as a promise pointing to the value of `x` and when it is
finally called upon, the value happens to be `20`.

One way to control when things are evaluated is to use the function
`force` which is defined simply as:

```{r}
force
```

So if we define `g2` as follows:

```{r}
g2 <- function(x, y = x) {
    force(y)
    x <- 2 * x
    y
}
g2(10)
```
we see that we get the expected answer.

## Scope

Scope pertains to where values of a symbol will be looked up.  The
actual look up happens when the function is run, according to certain
rules: the scoping rules. These rules specify how symbols are
evaluated by the evaluator. Let us examine this in the context of a
function definition.

```{r}
f <- function(x, y = 2) {
    w <- x^2 + y^2
    x + y + z + w
}
```
For a function, you have:

- __formal arguments__. These are the symbols in the argument
  list. When the function is invoked these have to be bound to some
  value (a promise object, as we told you). So they are bound symbols.
- We also have variables defined inside the function. These are
  _local_ variables and they are bound to values that they are
  assigned to.
- The other variables that are just used inside the function. These
  may or may not result in an error during evaluation if the variable
  cannot be found. So these variables are neither local (since they
  were not defined inside the function, i.e. not on the left hand side
  of an assignment) nor bound inside the function. You may say they
  are _free_.

In the above `f`, the variables `x` and `y` are bound to arguments, `w` is local,
and `z` is is unbound or free.

Recall that a `frame` is a set of symbol-value pairs and is evaluated
in some environment, the two together create a new environment which
is the function's environment.

When a function is called a new evaluation frame is created. The
function's enclosing environment is the environment active at the time
the function was created. This is made available to the function.

### Function call walkthrough

Let us see what happens as we call `f(1)`.

```{r, error = TRUE}
f(1)
```

- `x` gets bound to 1, `y` is bound to 2
- `w` is computed as `1^2 + 2^2` = `5`.
- Now to compute `x + y + z + w`, a value needs to be found for `z`.
So the search for `z` begins
- `z` is searched for in the environment of `f` which is
```{r}
environment(f)
```
the global environment.
- Since there is no `z` in the global environment, the error is
raised.
```{r}
ls(environment(f))  ## No z here?
```

So now let us make a `z`.
```{r}
z <- 2
f(1)
```
That works as `z` is now found.

A few more examples.

```{r, error = TRUE}
f <- function() {
  y
}
rm("y") ## Ensure we don't have y in workspace
f()  ## error since y not found
y <- 1
f()  ## ok, y is found a level up.
```

Now define `y` inside.

```{r}
f <- function() {
  y <- "y from inside"
  y
}
f()
y <- 1
f()  ## still same result
```

Return a function.

```{r, error = TRUE}
f <- function(x, y=2) {
    w <- x^2 + y^2
    x + y + z + w
}
rm(z)
f(1) ## fails
z <- 10
f(1) ## 1 + 2 + 10 + 5
rm(z)
f(1) ## fails again (dynamic behavior!)
```

I point this out because I know people have been surprised by this
behavior, particularly when you save your R workspace and load it
back.  Depending on what is in your workspace before you load back,
your results could be different.,

Now, this does not mean saving your R session workspace is a bad
idea. With large datasets, you often want to save the processing time.

The better thing to do is to save objects, rather than the entire
workspace. Note also, if your saved objects only make sense in the
context of a loaded package, unless that package is loaded again, you
could have very different behaviors.

It is vital for an R developer to code (functions) with careful
attention to scoping rules so that there are no harsh surprises.


### Another function call walkthrough

Consider:

```{r}
f2 <- function() {
  z <- 10
  g <- function(x, y = 2) {
      w <- x^2 + y^2
      x + y + z + w
  }
  g
}
```

Let us examine what is in each environment.

```{r}
ls(environment(f2))
```

```{r}
gFn <- f2()
environment(gFn)
parent.env(environment(gFn))
ls(environment(gFn))
```

Note that:
- environment of `gFn` is _not_ the global environment!
- the parent environment of `gFn` is in fact the global environment,
which is the environment of `f2`.
- `z` is _available in the environment of `gFn`, with value `10`,
`gFn(2)` will use that value.

```{r}
gFn(2)
```

We also have a `z` in the global environment, but that has no effect:
the scoping rules mean that the `z` in the environment of `gFn` will
be found and so there is no need to look it up in the global
environment, which is one level up!

_This behavior is often exploited preserve state in conjunction with
another assignment operator `<<-`. In fact, it is used in reference
based OOP classes in R.

## Other functions

Sometimes, it helps to have syntatic sugar.

One can perform set operations via infix operators. For example, here
are functions based on already existing functions: the functions
`union`, `setdiff`, etc. are already in R.

```{r}
`%+%` <- union
`%-%` <- setdiff
`%=%` <- setequal
`%*%` <- intersect
`%in%` <- is.element
```

We can then apply them in an infix fashion.

```{r}
A <- c("Cat", "Dog", "Parakeet")
B <- c("Dog", "Parakeet")
A %*% B  ## intersection
A %-% B  ## set difference
"Cat" %in% B  ## membership
```

Lest you get too excited about this and use it immoderately, beware
that you mask already defined operations. For example, `%*%` is used
for matrix multiplication.  But you can access the original definition:

```{r}
get("%*%", envir=baseenv())  ## so you can still use it
```

So such syntactic sugar should be used wisely.

>
> Syntactic sugar causes cancer of the semicolon
>           --- Alan Perlis, Epigrams of Programming
>

However, such sugar can be really useful if used judiciously.

### Replacement functions

Suppose you want to set the last element of a vector to some
value. One way is to use subset indexing.

```{r}
x <- 1:3
x[3] <- 10
x
```
We know that this is actually happening via functions, specifically
the `[` function.

```{r}
get('[')
```
So suppose you try to write your own function.

```{r}
setLast <- function(x, value) {
    x[length(x)] <- value
    invisible(x)
}
```

Let us try it.

```{r}
x
setLast(x, 100)
x
```

Nothing changed! Why?

For example, here is a function to replace the last element of a
vector.

```{r}
`setLast<-` <- function(x, value) {
    if (!is.vector(x)) stop("setLast: Expecting a vector")
    x[[length(x)]] <- value
    x
}

x <- list(1, 2, 3)
setLast(x) <- list(4, 5, 6)
x
```

The above is equivalent to `x <- ``setLast<-``(x, list(4, 5, 6))`

## Session Info
```{r}
sessionInfo()
```
