---
title: 23.2. Testing Code in Packages
output:
     slidy_presentation
---

```{r, echo = FALSE}
installIfNeeded <- function(packages, ...) {
    installedPackages <- installed.packages()[, 1]
    toInstall <- setdiff(packages, installedPackages)
    if (length(toInstall) > 0) install.packages(toInstall, ...)
}
installIfNeeded(c('RUnit', 'testthat', 'stringr'), repos = "https://cloud.r-project.org")
```


Wikipedia says:

> In computer programming, unit testing is a software testing method
> by which individual units of source code, sets of one or more computer
> program modules together with associated control data, usage
> procedures, and operating procedures, are tested to determine whether
> they are fit for use.


A test is merely of check of an actual result versus an expected
result.  In R, the obvious things to test are functions since the R
language has a functional flavor. So we'd usually be testing the
output of a function versus what was expected.

___

R has always had a mechanism for including tests in packages. The
package subdirectory `tests` is for test code. The basic idea is that
you place an R file which is run in an R session and the entire output
is compared with a saved file of expected output _that you provide as
well_. For example, in the `stats` package tests, there is a file
`glm.R`.

```{r, eval = FALSE}
## Example in which the fit for the null deviance fails to converge:
# https://stat.ethz.ch/pipermail/r-help/2012-May/313161.html
Y <- c(rep(0,35),1,2,0,6,8,16,43)
beta <- 42:1
cst <- lchoose(42, beta)
tau <- (beta^2)/2
fit <- glm(formula = Y ~ offset(cst) + beta + tau, family = poisson)
```

and also `glm.Rout.save`

```
R version 3.4.2 Patched (2017-11-17 r73741) -- "Short Summer"
Copyright (C) 2017 The R Foundation for Statistical Computing
Platform: x86_64-pc-linux-gnu (64-bit)

R is free software and comes with ABSOLUTELY NO WARRANTY.
You are welcome to redistribute it under certain conditions.
Type 'license()' or 'licence()' for distribution details.

R is a collaborative project with many contributors.
Type 'contributors()' for more information and
'citation()' on how to cite R or R packages in publications.

Type 'demo()' for some demos, 'help()' for on-line help, or
'help.start()' for an HTML browser interface to help.
Type 'q()' to quit R.

> ## Example in which the fit for the null deviance fails to converge:
> # https://stat.ethz.ch/pipermail/r-help/2012-May/313161.html
> Y <- c(rep(0,35),1,2,0,6,8,16,43)
> beta <- 42:1
> cst <- lchoose(42, beta)
> tau <- (beta^2)/2
> fit <- glm(formula = Y ~ offset(cst) + beta + tau, family = poisson)
Warning messages:
1: glm.fit: algorithm did not converge 
2: In glm(formula = Y ~ offset(cst) + beta + tau, family = poisson) :
  fitting to calculate the null deviance did not converge -- increase 'maxit'?
> 
> proc.time()
   user  system elapsed 
  0.189   0.034   0.203 
```

During the testing process, R simply runs `glm.R` and saves the output
to `glm.Rout` and compares that to `glm.Rout.save`.  Differences are
reported, but they don't cause an error.

___

Obviously, this kind of testing can only go so far.

- No test of R structures
- Only approximate comparisons
- Accuracy problems.

So a better testing mechanism is needed.

##  Session Information
```{r}
sessionInfo()
```









