---
title: 23.3. `RUnit` Testing
output:
     slidy_presentation
---

__NOTE__ I provide this merely for completeness, so that others who
come from a Java world may find it easy to recognize the testing
framework. I highly recommend `testthat`.

___

`RUnit` provides a simple mechanism for testing output of R
expressions. Example from vignette. Consider

```{r}
library(RUnit)
c2f <- function(c) return(9/5 * c + 32)
```

A corresponding test function could look like this:

```{r}
test.c2f <- function() {
  checkEquals(c2f(0), 32, " Checking 0C = 32F")
  checkEquals(c2f(10), 50, " Checking 10C = 50F")
  checkException(c2f("xx"), " Checking exception with non numeric arg")
}
```

The default naming convention for test functions is `test.` as in
`JUnit` which is the inspiration for `RUnit`.

___

One normally creates a suite of such tests in a file and runs
them. The function `defineTestSuite` helps.

```{r}
testsuite.c2f <- defineTestSuite(name = "c2f",
                                 dirs = file.path(path.package(package="RUnit"), "examples"),
                                 testFileRegexp = "^runit.+\\.R",
                                 testFuncRegexp = "^test.+",
                                 rngKind = "Marsaglia-Multicarry",
                                 rngNormalKind = "Kinderman-Ramage")
```
Then all you do is run the test suite:

```{r}
testResult <- runTestSuite(testsuite.c2f)
printTextProtocol(testResult)
```

You can also run a single file:

```{r, eval = FALSE}
runTestFile(file.path(system.file(package="RUnit"),
                      "examples/runitc2f.r"))
```
___

###  23.3.1.1 `RUnit` equality checks

What kinds of equality checks are available in `RUnit`?

```{r, eval=FALSE}
checkEquals(target, current, msg, tolerance = .Machine$double.eps^0.5,mcheckNames = TRUE, ...)
checkEqualsNumeric(target, current, msg, tolerance = .Machine$double.eps^0.5, ...)
checkIdentical(target, current, msg)
checkTrue(expr, msg)
checkException(expr, msg, silent = getOption("RUnit")$silent)
```

___

### 23.3.2. Example use in `sp23Design`

A fairly involved package for group sequential seamless Phase II-III
Clinical trials. Include interim looks, a number of parameters etc.

```{r, eval=FALSE}
###################################################
### code chunk number 38: Case-A-unit-test
###################################################
test.caseA <- function() {
  trialParameters <- list(minimumNumberOfEvents = 20,
                          minimumIncreaseInV = 0.2,
                          numberRecruitedEachYear = c(80, 120, 160, 160),
                          followupTime = 3,
                          adminCensoringTime = 7,
                          interimLookTime = c(1, 2, 3, 5, 7),
                          type1ErrorForResponse = 0.05,
                          type2ErrorForResponse = 0.01,
                          glrBoundarySidedness = "one", # one sided or two-sided
                          type1Error = 0.05,
                          type2Error = 0.10,
                          epsType1 = 1/3,
                          epsType2 = 1/3)
  trueParameters <- list(p0 = 0.3,
                         p1 = 0.3,
                         pdiffHyp=0.3,
                         theta = list(
                           alpha = 0,
                           beta = 0,
                           gamma = 0),
                         baselineLambda = 0.35,
                         etaHyp = 0.25)
  rngSeed <- 9872831
  sp23Design <- generateSP23Design(trueParameters, trialParameters)
  trialHistory <- exploreSP23Design(sp23Design, numberOfSimulations=25, rngSeed=rngSeed)

  result <- analyzeSP23Design(sp23Design, trialHistory)$designSummary

  checkEquals(result[["numberOfTimesH0RIsRejectedAtFirstLook"]], 0)
  checkEquals(result[["numberOfTimesH0RIsRejected"]], 0)
  checkEquals(result[["numberOfTimesStoppedForFutility"]], 25)
  checkEquals(result[["numberOfTimesH0SIsAccepted"]], 0)
  checkEquals(result[["numberOfTimesH0SIsRejected"]], 0)
  checkEquals(result[["numberOfTimesFutilityDecidedAtLastLook"]], 0)
  checkEquals(result[["numberOfTimesTrialEndedAtLook"]],
  {
      out <- c(9, 13, 3, 0, 0)
      names(out) <- trialParameters$interimLookTime;
      out}
  )
}
```

___

### 23.3.3. `RUnit` summary

`RUnit` does not directly integrate with the `R CMD check` mechanism.
The package `svUnit` was written to address this and other
issues. It also produces nicer summaries for use in web pages etc.

___

## Session Info

```{r}
sessionInfo()
```
