---
title: 23.5. Global Variables in Packages
output:
     slidy_presentation
---

One should really try very hard to avoid global variables in
packages. This is because global variables make code harder to reason
with; they can can introduce subtle race conditions and therefore
bugs. Use of global variables in multi-threaded languages will usually
cause trouble.

___

R is single threaded, that is, it processes one task at a time.
Together with R's scoping rules, this allows one to get away with
global variables. In fact, many packages work this way. However,
global variables are not needed as often as one might think.  Some
defensible uses of global variables:

- For setting convergence thresholds or numerical tolerances say in a
numerical routine
- For setting a maximum number of iterations for a calculation that
might not terminate
- For setting default plotting options etc.

## 23.5.1. Example

Consider a test package named `testpack1`. To keep things simple, I have
introduced a new function:

```{r, eval = FALSE}
closeEnough <-  function(x, y) {
    ## check if two numbers are close enough, within testpack1_defaults$tol
    abs(x - y) <= testpack1_defaults$tol 
}
```

Here we have introduced a variable called `testpack1_defaults`, a list
containing parameter `tol` to judge closeness. You can imagine a
situation where it might be good to let the user set this option,
depending on the user's notion of closeness. For example, a user might
desire `tol = 1e-2` even though we set a value of `tol = 1e-3` as a
default.

We may have other values too, such as a default number of iterations
for some numerical routine (`testpack1_defaults$max_iter`)

How and where to define it?

## 23.5.2 Approach 1: Use `options()`

Recall that `options()` is a long list of settings that controls how R
works. We can use it to store any value we wish. 

In a package, you can do this in any file at the top level; we do it
in a file called `globals.R`.

```{r, eval = FALSE}
##
## contents of file testpack1/R/globals.R
#' Function to set defaults
#' @param tol the tolerance, default 1e-3
#' @param max_iter the maximum number of iterations, default 25
#' @return a list of option values invisibly
#' @examples
#' setDefaults(tol = 1e-5)
#' getOption("testpack1_defaults")
#' @export
setDefaults <- function(tol = 1e-3, max_iter = 25) {
    result  <- as.list(environment())
    options(testpack1_defaults = result)
    invisible(result)
}

## When package is loaded, ensure that defaults are set
.onLoad  <- function(libname, pkgname) {
    setDefaults()
}

## When package is loaded, ensure that defaults are unset
.onUnload  <- function(libname, pkgname) {
    options(testpack1_defaults = NULL)
}
```

In the above, we have a function `setDefaults` that has all the
default parameters we want as arguments. 

The variable `result` gets set to a list of all variables in the
current environment; recall that `environment()` returns the current
environment at that point in the function and so will have all the
passed parameters.

We then set the options and return the result invisibly.

The `.onLoad` and `.unLoad` package hook functions are _crucial_! What
are those? 

- When a package is _loaded_, `.onLoad` is called, if you define
it. The `libname` is the directory where the package defining the
namespace was found. The `pkgname` is the character string giving the
package name. 

- When a package is _unloaded_, the `.unLoad` function is called, once
  again with the same arguments as above.

Without `.onLoad`, the defaults would never be set when the package is
loaded!

___

We can check how the `.onLoad` and `.unLoad` functions work by simply
loading `testpack`.

First, let us look at our search path.

```{r}
search()
```
We don't have `testpack1` loaded. So let us check on option
`testpack_defaults`

```{r}
getOption("testpack1_defaults")
```

Let us install `testpack1` and load it.

```{r}
install.packages("testpack1_0.1.tar.gz", type = "source", repos = NULL)
library(testpack1)
getOption("testpack1_defaults")
```

Now `testpack1_defaults` became available. Let us detach it,
i.e. remove from search path.

```{r}
detach(2)
getOption("testpack1_defaults")
```

Still `testpack1_defaults` is available, because although we removed
`testpack1`  from search position 2, it is still loaded!

```{r}
loadedNamespaces()
```

So let us unload the namespace and check.

```{r}
unloadNamespace("testpack1")
getOption("testpack1_defaults")
```

Indeed the option is now gone.


## 23.5.3 Drawback of `options()`

Many package use option variables to operate. However, they are not
very safe. For, consider the following.

```{r}
library(testpack1)
example(closeEnough)
```

But now let's be mischievous. 

```{r}
## Clobber the testpack1_defaults option
options(testpack1_defaults = NULL)
example(closeEnough)
```

Although the code does not throw a error, it is wrong!

The root cause: we have more than one way of setting options which
means we do not have full control of the variable!

Another drawback is that the `setDefaults` function is a bit rigid: we
cannot change only one value of the parameter while keeping the
previous value for others. 

## 23.5.4 Approach 2: Using own list (Recommended!)

Maintain your own list of options.

This is outlined in `testpack2` in the file `testpack2/R/globals.R`.

```{r, eval = FALSE}
##
## Global variable for package testpack
##
testpack2_defaults  <- list(tol = 1e-3, max_iter = 25)

#' Function to set defaults
#' @param tol the tolerance, default 1e-3
#' @param max_iter the maximum number of iterations, default 25
#' @return a list of default values invisibly
#' @examples
#' (resetDefaults())
#' (resetDefaults(tol = 1e-5))
#' @export
resetDefaults <- function(tol = 1e-3, max_iter = 25) {
    testpack2_defaults  <- as.list(environment())
    invisible(testpack2_defaults)
}

#' Set the default param name to a value
#'
#' @param name the parameter name
#' @param value the value to assign
#' @return the new list of defaults invisibly
#' @examples
#' (setDefaults("tol", 1e-5))
#' (setDefaults("max_iter", 100))
#' @export
setDefaults <- function(name, value) {
    ## check if param name is known!
    if (!(name %in% names(testpack2_defaults))) {
        stop(sprintf("Bad parameter name %s", name))
    }
    testpack2_defaults[[name]]  <- value
    invisible(testpack2_defaults)
}

```

___

Again, we start by installing the package and make sure this new
approach works.

```{r}
install.packages("testpack2_0.1.tar.gz", type = "source", repos = NULL)
library(testpack2)
testpack2::closeEnough(1.01, 1.02) ## FALSE per default tolerance
setDefaults("tol", .1) ##
testpack2::closeEnough(1.01, 1.02) ## TRUE per set tolerance
```

## 23.5.3. A more involved example

In the `disctomp` package, we have several `shiny` apps that ease the
task of defining a computation, uploading it to an `opencpu` server,
etc.

Consider for example, the task of defining a computation. One does
this via `defineNewComputation()`.  The task of this function is to
first let a user choose a computation from a list of available ones
and then, based on the choice get the appropriate inputs via another
shiny app tailored for that choice. For example, with `distcomp` one
chooses a type of computation first. Then based on that computation,
you get further parameters you _specific to that choice.__

```{r, eval = FALSE}
library(distcomp)
defineNewComputation()
```

___

The easiest way to do this in R is via environments in any one of your
R files in the package at the top level.

```{r, eval = FALSE}
pkgGlobals <- new.env( parent=emptyenv())
```

You need the `parent` argument to be the empty environment because you
want the _enclosing environment_ to be empty. Typically, you'll be
looking for variables in this environment and you don't want to find
something of the same name in some enclosing environment.

Then you can set up `set` and `get` methods to pick of variables you
want.

```{r}
pkgSetGlobal <- function(name, value) {
  pkgGlobals[[name]] <- value
}
pkgGetGlobal <- function(name) {
  pkgGlobals[[name]]
}
```

Note that the return value in the latter might be `NULL`.

In the `distcomp` application, what happens is

1. the global variable gets set for the type of computation
2. the currently running `shiny` app is stopped, and returns the
   value of the chosen computation
3. A new appropriate `shiny` app is fired up based on the chosen
   computation type and makes use of the global variables

## 23.5.4. Caution

It is easy to get carried away with this in `shiny` because you may
have several UI elements react to certain global values being set. To
see this, consider what happens when a UI button is pressed.

Suppose that when the button is pressed, you update a global variable
and then have several other UI elements change their
appearance. However, you cannot guarantee the order in which

1. the global variable will be updated
2. those UI elements will be updated.

So you should really think very carefully about using global
variables.

## Session Information
```{r}
sessionInfo()
```



