---
title: 23.6. Those Pesky Notes
output:
     slidy_presentation
---

Consider the package we created in Assignment 2. You had a plotting
function `plotTzbyOS` using `ggplot2`.

Recall the function definition.

```{r, eval = FALSE}
plotTzByOS <- function(data, color = c("blue", "#E69F00")) {
    tzAndOs <- tibble(tz = unlist(lapply(data, function(x) x$tz)),
                      agent = unlist(lapply(data, function(x) x$a))) %>%
        mutate(tz = ifelse(tz == "", "Unknown", tz),
               os = ifelse(grepl("Windows", agent), "Windows", "Nonwindows")) %>%
        select(tz, os)
    ... ## Stuff omitted for brevity
}
```

___

You might have also noticed that in the file `stat290.ass2.R` we had
the following.

```{r, eval = FALSE}
utils::globalVariables(names = c("os",
                                 "Nonwindows",
                                 "Windows",
                                 "value",
                                 "agent",
                                 "count",
                                 "tz"))

```

Ordinarily, if you did not that incantation above---comment it out,
for example---and you ran `R CMD check` on it, you would get
complaints as shown below.

```
* checking dependencies in R code ... OK
* checking S3 generic/method consistency ... OK
* checking replacement functions ... OK
* checking foreign function calls ... OK
* checking R code for possible problems ... NOTE
plotTzByOS: no visible binding for global variable ‘tz’
plotTzByOS: no visible binding for global variable ‘agent’
plotTzByOS: no visible binding for global variable ‘os’
plotTzByOS: no visible binding for global variable ‘count’
Undefined global functions or variables:
  agent count os tz
* checking Rd files ... OK
* checking Rd metadata ... OK
* checking Rd cross-references ... OK
...
* DONE
Status: 1 NOTE
```

___

Let's examine what that complaint is about.

R is asking: 

Dude, where are you getting the parameter `tz` and `agent` in the call
to `mutate`? I see no variables named `tz` or `agent`. Are you sure
that is not something you overlooked?  

Similarly, `os` in the `select` call and others further down.

___

The root cause is that we are using `dplyr` constructs; packages
like `dplyr`, `ggplot2` and `magrittr` do all kinds of non-standard
and deferred evaluation. 

In this example, `dplyr` knows to evaluate `tz` and `agent` in the
context of the tibble constructed earlier in the pipeline. In other
words, when the code is executed, those variables will be available
and everything will work fine. (Well, at least we assume so!)

But that is not possible for R to detect via a code analysis which is
what it does during `R CMD check`. The reality is that there is no
guarantee that the code will run correctly! For example, nothing stops
a user from passing a data frame that has _no_ variable called
`os`. In that case, the code will fail.

To paraphrase Djikstra: _Running the code can only  detect presence of
errors, not their absence_.

___

## 23.6.1. The solution

The `R-exts` manual suggest you use the following construct in your
package to avoid _known to be false_ `NOTE`s.  It is important not to
abuse this. 

```{r, eval = FALSE}
utils::globalVariables(names = c("os",
                                 "Nonwindows",
                                 "Windows",
                                 "value",
                                 "agent",
                                 "count",
                                 "tz"))

```

Then everything works.  Note also that here I have included more
variables than necessary (`Nonwindows`, `Windows`, `value`) and it is
easy to make errors.

It is important not to abuse this facility to make even legitimate
`NOTE`s go away!

___

## Session Information
```{r}
sessionInfo()
```



