---
title: "28.1. Parallel Random Number Generation"
output:
      slidy_presentation
---

In R, the random number generator gets seeded, on demand, using the
clock time and a process id. The state of the random number generator
is saved in a structure `.Random.seed`. In a fresh R session, this
structure is not to be found.

```{r}
exists(".Random.seed")
```

Upon invoking any R function that uses the random number generator,
the structure gets created.

```{r}
rn <- rnorm(1)
str(.Random.seed)
```

___

When using parallel or cluster computing, a worker process can get the
same random number state simply because a workspace containing
`.Random.seed` was restored or the same random number generator was
used before the process forked.

```{r}
runif(1)
save.image("session.RData")
runif(1) ## Note this value
load("session.RData")
runif(1) ## Should be same as the last one
```

___

In other cases, each worker process may get its own seed, depending on
 the situation.
 
This means you have a very good chance of not being able to reproduce
any of your results/errors.

Therefore, when we use random numbers in your parallel computations,
the following are desiderata.

1. We would like to ensure reproducibility. Otherwise, we'll never be
able to debug odd results or programs

2. We want to ensure that the worker nodes/processes to use
independent random sequence streams with _large periods and good
theoretical properties_. 

Setting a seed in each worker in a reproducible way will work, but the
random number streams generated this way may get into step because
although, the seeds may be different, but the generator is the
same. Not all generators were designed explicitly not to get into
step.

3. Since coming up with different seeds for each process can become a
chore, we would also like to initialize them with a small number of
seeds. Like one number (`2893` for example), preferably in a master
process and then have a standard way of working with the random number
streams in the worker processes.

4. You would like to jump ahead in streams. This is for efficiency as
   well as debugging: it means that you can easily jump to problematic
   simulations.

5. Finally, you would like the entire computation to be reproducible.

___

The `parallel` package contains an implementation of the ideas of
L’Ecuyer et al. (2002).

- It uses a single RNG and make streams with seeds $2^{127}$ steps
apart in the random number stream (which has period approximately
$2^{191}$). This generator is that it has a fairly long period with a
small seed (6 integers), and unlike R’s default "Mersenne-Twister"
RNG, it is simple to advance the seed by a fixed number of steps.

## 25.3.1. Examples

Since the key issues have to do with the random number generation and
therefore, we will take work with simple examples involving uniform
random number generation. (All random variables rely on a uniform
random number generator!)

Typically, you would set the seed for the random number generator and
generate some sequence of random variables.

___

Let us begin with a simple R session. 

```{r}
set.seed(123)
seqRNG <- runif(15)
print(seqRNG)
```

Now in a parallel environment, let us try the same thing. We generate 3
sets of 5 after resetting the seed.

```{r}
library(parallel)
cl <- makeCluster(3)
set.seed(123)
rngs <- clusterApply(cl, 1:3, function(x) runif(5))
print(rngs)
```

___

Ok, we got different results back from each worker, but at least do we
get the same result back from all nodes if we repeat the process?

```{r}
set.seed(123)
rngs <- clusterApply(cl, 1:3, function(x) runif(5))
print(rngs)
```

No! That is because, we the above code reset the seed only on the
master and that had no effect on the workers which are running
independent R processes. 

___

So we fix that by setting the seed on all workers.

```{r}
clusterEvalQ(cl, set.seed(123))
rngs <- clusterApply(cl, 1:3, function(x) runif(5))
## Produces same sequence on all of them
print(rngs)
clusterEvalQ(cl, set.seed(123))
rngs <- clusterApply(cl, 1:3, function(x) runif(5))
## Produces same sequence on all of them
print(rngs)
stopCluster(cl)
```

Ok, that did it; we have reproducibility but that is useless if you
are doing simulations: all workers are doing the same thing because
they are all using the same stream of random numbers.

___

The same also happens with the unified approach we discussed using
`foreach` and `%dopar%`. For completeness, here it is.

```{r}
library(foreach)
library(doParallel)
cl <- makeCluster(3)
registerDoParallel(cl)
set.seed(123)
res <- foreach(i = 1:3) %dopar% { runif(5) }
```

```{r}
set.seed(123)
res2 <- foreach(i = 1:3) %dopar% { runif(5) }
identical(res, res2)
print(res)
stopCluster(cl)
```

```{r}
library(doParallel)
cl <- makeCluster(3)
registerDoParallel(cl)

set.seed(123)
res <- foreach(i = 1:3) %dopar% { runif(5) }

set.seed(123)
res2 <- foreach(i = 1:3) %dopar% { runif(5) }
identical(res, res2)
```

___

Again, as before we got non-identical results, which can be fixed by
setting seeds in the worker nodes, but we again get useless (repeated)
simulation results.

```{r}
clusterEvalQ(cl, set.seed(123))
res <- foreach(i = 1:3) %dopar% { runif(5) }
clusterEvalQ(cl, set.seed(123))
res2 <- foreach(i = 1:3) %dopar% { runif(5) }
identical(res, res2)
stopCluster(cl)
```


## 28.1.2. Setting the default RNG kind

We need to understand how random number generation in R works. By
default, R uses the Mersenne-Twister generator.  This is documented in
the help page for `set.seed`. Let us take a look at the help for
`set.seed`.

Note that `set.seed` is the recommended way of seeding random number
generators in R. Also note that the variable `.Random.seed` in the
global environment stores the _state_ of the random number generator.

When a brand new R session is started, there is no variable
`.Random.seed`. 

Once you make a call to any of the random number generation functions,
R looks for `.Random.seed` and if it does not find it, generates
one. This is done by a call to `set.seed` using some combination of
time and date and other system parameters. Since this is not something
you can control, the only way to guarantee reproducibility is by
seeding the generators yourself.

___

What is the default generator R provides?

```{r}
RNGkind()
```

For parallel computations, we want to use L`Ecuyer's Combined Multiple
Recursive Generator (CMRG) which provides facilities for multiple
streams because of desiderata 1. This generator is defined by the
following recursion:

$$
  x_n = 1403580 \times x_{n-2} - 810728 \times x_{n-3} \mod{(2^{32} -
  209)}
  $$

$$
y_n = 527612 \times y_{n-1} - 1370589 \times y_{n-3} \mod{(2^{32} -
22853)}
$$

$$
z_n = (x_n - y_n) \mod{4294967087}
$$

$$
u_n = z_n/4294967088 \mbox{ unless $z_n = 0$}
$$

This recursion has to be seeded with a 6-vector:

$$
(x_{n-3}, x_{n-2}, x_{n-1}, y_{n-3}, y_{n-2}, y_{n-1}).
$$

___

The recursion for each of $x_n$ and $y_n$ can have pre-computed
coefficients for $k$ steps ahead.  For $k = 2^{127}$, the seed is
advanced by $k$ steps by

```{r, eval = FALSE}
.Random.seed <- nextRNGStream(.Random.seed)
```

Furthermore, sub-streams can be chosen starting from seeds $2^{76}$
steps apart using `nextRNGSubStream()` which advances to the next
substream.

So, we set the default generator to be L'Ecuyer's CMRG on all workers.

```{r}
cl <- makeCluster(3)
clusterSetRNGStream(cl, 12345)
str(clusterApply(cl, 1:3, function(x) list(rngKind = RNGkind(), state = .Random.seed)))
```

___

The function `clusterSetRNGStream` selects the L'Ecuyer's CMRG and
distributes streams to the members of a cluster. It takes _one_ seed as an
argument (`12345`).

We can now get some random draws.  


```{r}
crngs <- clusterApply(cl, 1:3, function(x) runif(5))
print(crngs)
```

___

Now, if we set the stream again, we should get the same result as above.

```{r}
clusterSetRNGStream(cl, 12345)
clusterApply(cl, 1:3, function(x) RNGkind())
crngs2 <- clusterApply(cl, 1:3, function(x) runif(5))
## Should be same results as before
print(crngs2)

identical(crngs, crngs2)

stopCluster(cl)
```

## 28.1.3. The package `doRNG`

The `doRNG` package does the above in a principled way. You use it as
follows.

```{r}
library(doRNG)
cl <- makeCluster(3)
registerDoParallel(cl)

set.seed(123)
res <- foreach(i = 1:3) %dorng% { runif(5) }
set.seed(123)
res2 <- foreach(i = 1:3) %dorng% { runif(5) }
identical(res, res2)
stopCluster(cl)
```

___

How does `doRNG` work?

For a loop with N iterations, the `%dorng%` operator internally performs
the following tasks:

1. Generate a sequence of random seeds $(S_i)_{1 \leq i \leq N}$ for
   the R random number generator L'Ecuyer-CMRG using the function
   `nextRNGStream` from the `parallel` package, which ensure the
   different RNG streams are statistically independent

2. Modify the loop's R expression so that the random number generator
   is set to L'Ecuyer-CMRG at the beginning of each iteration, and
   is seeded with consecutive seeds in $(S_n)$: iteration $i$ is seeded
   with $S_i$, $1 \leq i \leq N$

3. Call the standard `%dopar%` operator, which in turn calls the
   relevant (i.e. registered) `foreach` parallel backend

4. Store the whole sequence of random seeds as an attribute in the
   result object.

The last step is important because you now have a way of getting those
seeds and reproducing even parts of the results later if necessary by
resetting the seeds.

```{r}
attr(res, "rng")
```


## 28.1.4. Session Info

```{r}
sessionInfo()
```
