---
title: "10.0. Revised R Package"
output:
     slidy_presentation
---

```{r, echo = FALSE}
## Install a package if not already installed
installIfNeeded <- function(packages, ...) {
    toInstall <- packages[!(packages %in% installed.packages()[, 1])]
    if (length(toInstall) > 0) {
        install.packages(toInstall, repos = "https://cloud.r-project.org")
    }
}

## Ensure packages exist and activate them
needPackages <- function(packages) {
    installIfNeeded(packages)
    for (x in packages) {
        library(x, character.only = TRUE)
    }
}
needPackages("devtools")
```

## 10.0.1. Revising our package again

We are now going to make more changes to our revised package.

## 10.0.1. Add a data set

We will use the bitly dataset. Assuming that you are in the correct
package directory, you can specify a data set as part of your package
as follows. 

```{r, eval = FALSE}
bitly  <- readRDS("bitly.RDS")
usethis::use_data(bitly)
```

Notice what changes this makes to the package source files and folders.

## 10.0.2. Adding a plotting function

Let us add a very simple scatterplot function using the `ggplot2`
package. We first announce our intention by ensuring that the
`DESCRIPTION` file, which indicates the imports and dependencies is
appropriately modified. 

```{r, eval = FALSE}
usethis::use_package("ggplot2", type = "Imports")
```

___

Let us add a very simple scatterplot function by putting the following
code in `basicplot.R` in the R subdirectory of the package.

```{r, eval = FALSE}
simple_plot  <- function(x, y) {
    ggplot() +
        geom_point(mapping = aes(x = x, y = y))
}
```

___

We would like to make this available to our package users.  So we
better document it and export it!

```{r, eval = FALSE}
#' A simple scatterplot using ggplot2
#' 
#' @param x the abscissa values
#' @param y the ordinate values
#' @return a ggplot object
#' @examples
#' simple_plot(mtcars$wt, mtcars$mpg)
#' @export simple_plot
simple_plot  <- function(x, y) {
    ggplot2::ggplot() +
        ggplot2::geom_point(mapping = ggplot2::aes(x = x, y = y))
}
```

___

However, that documentation is not sufficient. We have not told R that
this plotting function relies on specific functions from `ggplot2`!

So we actually need to indicate this via:

```{r, eval = FALSE}
#' A simple scatterplot using ggplot2
#' 
#' @param x the abscissa values
#' @param y the ordinate values
#' @return a ggplot object
#'
#' @importFrom ggplot2 ggplot geom_point aes
#' @examples
#' simple_plot(mtcars$wt, mtcars$mpg)
#' @export simple_plot
simple_plot  <- function(x, y) {
    ggplot2::ggplot() +
        ggplot2::geom_point(mapping = ggplot2::aes(x = x, y = y))
}
```

## 10.0.3. Build and check

Now we can `devtools::document()`, `devtools::build()` and
`devtools::check()`.

I do these via command lines. 

## 10.0.4. Important Note

I want to strongly iterate that it is not uncommon to run into issues
that cannot always be solved neatly via `devtools` or Rstudio IDE.

_You must be willing to get your hands dirty._

This means completely mastering the organization of R packages.  It is
worth the effort!


## Session Info
```{r}
sessionInfo()
```
