---
title: 10.1. R Graphics
output:
     slidy_presentation
---
```{r, echo = FALSE, results = 'hide', message = FALSE}
## Install a package if not already installed
installIfNeeded <- function(packages, ...) {
    toInstall <- packages[!(packages %in% installed.packages()[, 1])]
    if (length(toInstall) > 0) {
        install.packages(toInstall, repos = "https://cloud.r-project.org")
    }
}

## Ensure packages exist and activate them
needPackages <- function(packages) {
    installIfNeeded(packages)
    for (x in packages) {
        library(x, character.only = TRUE)
    }
}
needPackages(c("tidyverse", "png"))
```


The S/R system was designed with publication quality graphics from the
start. However, over time, new packages and facilities have been added
to enhance the system.

## 10.1.1. Graphics Structure

```{r, echo = FALSE}
library(grid)
img <- readPNG("gorg.png")
grid.raster(img)
```
[_Source_: R Graphics by Paul Murrell.]


## 10.1.2. Base Graphics

To see what base graphics can do, try `demo(graphics)`.

Common functions are:

- `plot`, `lines`, `points` for common
    $x-y$ plots, adding lines, points
- `axes` for drawing axes, `box` for rectangular box
- `text`,  `mtext` for placement of text
- `pairs` for scatterplot matrices, `coplot` for
    conditioning plots
- Some interactivity: `identify`, `locator`
- Higher level plots: `rug`,
    `dotchart`, `hist`, `barplot`,
    `boxplot`, `qqplot`, `qqnorm`, `image`


As we already saw, many R objects have default plotting methods.
Every aspect of plotting can be controlled.

## 10.1.3. Graphics Devices

The graphics packages use low-level device software (in the `grDevices`
  package).

Eventual output is via screen software or document format.

Default device plots to the screen using standard system graphics: `quartz`
  on Mac, `x11` in Linux.

Many graphics devices are available and you can create devices that
write to a file (e.g., `pdf` for pdf output, `png` for png, `jpeg` for
jpeg, etc.).

To save a graphic that you have created, say in a `pdf` format, you
use the corresponding graphic device, thus.

```{r, eval = FALSE}
pdf(file = "plot.pdf")
plot(x = 1:10, y = 1:10)
dev.off()
```
 
## 10.1.4. Graphics parameters

The `par` can be used to set graphics parameters, the most complicated
function in R by some measure.

Things you can control.

- Layout, size, fonts
- Plot symbols, fill, line types
- Labels, annotations, text placement
- Axes, tickmarks
- Interactivity (`ask`)
- Colors
- Transparency (alpha blending) on some devices
- Support for plotting math symbols (`plotmath`

Grid equivalents are specified via `gpar`.

## 10.1.5. Example

Many objects in R have a default plotting method. What does that mean? 

It means that you can pass that object to the `plot` function and the
appropriate plot will be generated. Consider the following dataset.

```{r}
library(graphics)
data(pressure)
class(pressure)
```

Let's plot it.

```{r}
plot(pressure)
```
Note how the plot function _knew_ what to do with the data frame
`pressure`. Let's examine how this actually happens.

___

```{r}
plot
```
shows that `plot` is a function in the `graphics` package. 

A _generic function_ is one that behaves differently depending on the
class of its arguments. 

Contrast with standard OOP where objects have methods and so a
similarly named method `foo.plot()` and `bar.plot()` can potentially
generate different types of plots for objects `foo` and `bar`.

A generic function is a functional way of achieving the same effect:
`plot.foo` and `plot.bar` are functions that can be called if `foo` or
`bar` are (respectively) passed to the `plot` function. 

Let's see what else `graphics` provides for plotting.

```{r}
str(pressure)
class(pressure)
```

```{r}
ls(as.environment("package:graphics"), pattern = "^plot")
```

Hmmm, we don't see a `plot.data.frame` which is what we were expecting
to see.  But such functions may yet be there and not visible for
general use.

One way to find such functions is by looking at the _methods_ for a
generic function. These methods are the functions that actually do the
real work.

```{r}
methods(plot, data.frame)
```

Notice that `plot.data.frame` is not visible, i.e. exported. However,
we can still examine it using triple colon access:

```{r}
graphics:::plot.data.frame
```

__NOTE__ You will never use the triple colon functions in normal
scripts and packages because those are not exported for a reason. In
other words, you will only use the documented behaviors of a package
to ensure your code always works!

So `plot(pressure)`, since it is generic, did the following:

- Determined the class of `pressure` and invoked the _method_ plot for
data.frame `graphics:::plot.data.frame(pressure)` since `pressure` is a data
frame
- `graphics:::plot.data.frame` found that `pressure` had exactly 2
  elements and therefore called the _internal_ `plot2` function, which
  invoked `plot` one more time.
- The latter invocation tried to see if there was a method for
  `plot(x, y)`  where `x` and `y` are numeric since
  `class(pressure[[1]])` is `numeric`. Finding no such method, it
  invoked `plot.default(pressure[[1]], pressure[[2]])`


## 10.1.6. Plot layouts

You can control how plots are laid out on a rectangular grid. For
example, for two side-by-side plots:

```{r}
opar <- par(mfrow=c(1,2)) ## 1 row, 2 cols
plot(1:10)
plot(11:20)
par(opar)
```

__Warning__ `par()` is one of the most complicated functions in base
plotting. It has so many (interdependent) options that you will have
to examine it in detail to get the effect you desire. Since it
controls _global_ plotting options, it has side effects.

Also note the general style of programming above that changes plotting
parameters and retains the old settings in `opar` that you can
restore, thereby avoiding unintended consequences for subsequent
plots.


## 10.1.7. Building up plots

Every part of plotting can be controlled. So you can build up plots
step by step.

```{r, echo = FALSE}
library(grid)
img <- readPNG("xy.png")
grid.raster(img)
```


[_Source_: R Graphics by Paul Murrell.]

- Create an empty plot: `plot.new()`
- Set up the coordinate system:
`plot.window(range(pressure$temperature), range(pressure$pressure))`
- Plot the xy points: `plot.xy(pressure, type="p")`
- Add lines to join the points also
- Draw a box around it: `box()`
- Now the axes, horizontal, vertical and top: `axis(1);axis(2); axis(3)`
- Add the title: `title(xlab=names(pressure)[1], ylab=names(pressure)[2])`
```

Put it all together:

```{r}
plot.new()
plot.window(range(pressure$temperature), range(pressure$pressure))
plot.xy(pressure, type = "p")
lines(pressure)
box()
axis(1); axis(2); axis(3)
title(xlab = names(pressure)[1], ylab = names(pressure)[2])
```

Compare with the above `plot(pressure)`.

---

## 10.1.8. Exporting plots

```{r}
pdf(file = "plot.pdf", width = 5, height = 3)
plot.new()
plot.window(range(pressure$temperature), range(pressure$pressure))
plot.xy(pressure, type = "p")
lines(pressure)
box()
axis(1); axis(2); axis(3)
title(xlab = names(pressure)[1], ylab = names(pressure)[2])
dev.off()
```

---

## 10.1.9. What to use?

- Base graphics is very capable, but has aged a bit. Cannot completely
  avoid it since it is used in many packages via `plot` methods.

- The `grid` by Paul Murrell provides a different graphics engine that
  can take the place of base graphics. It is clear that this is the
  modern way to go since you can deal with objects and manipulate
  various phases of the plotting process. However, this is _too
  low-level_ for users.

- The `lattice` by Deepayan Sarkar, implements some of the ideas of
  Bill Cleveland's Trellis graphics in R. Uses the `grid` package to
  render the plots. If you are very familiar with base graphics, it
  might be easier to find lattice plot equivalents.

- The `ggplot2` is recommended and widely popular. It is also based on
  `grid` (so good!). However, because it can take a while to really
  master it.

---

## 10.1.10. Session Info

```{r}
sessionInfo()
```
