---
title: 10.2. Lattice Graphics
output:
     slidy_presentation
---

Lattice is very capable and improves on traditional base graphics. The
design goals were:

- Provide tools to apply general rules for creating more
    effective graphs
- Use as much of the available space as possible
- Force direct comparsion by superposition (grouping) when
    possible
- Encourage comparison when juxtaposing (conditioning): use
    common axes, add common reference objects such as grids

___

## 10.2.1. Lattice model

- A display is made up of various elements
- The defaults are coordinated to provide meaningful results, but each
    element can be controlled by the user independently of the others

The main elements are:

- the primary (panel) display
- axis annotation
- strip annotation (describing the conditioning process)
- legends (typically describing the grouping

Additional arguments to the high-level functions are used to activate
common variants, and full flexibility is allowed through arbitrary
user-deﬁned functions (e.g. controlling the primary display through
panel functions)

Each density plot is in an area called _a panel_. Panel sizes
determined automatically. Axes annotated with tick marks and labels
only along boundaries saving space. Strips atop each panel display
info

A _formula_ describes the content of the plots.

___

## 10.2.2. The formula argument

The first argument to many lattice functions is a formula argument
describing variables to plot. A trellis formula looks like

```{r, eval = FALSE}
     y ~ x | a * b
```

The `~` is what makes it a formula. The vertical bar denotes denotes
conditioning and may be omitted.

So `a` and `b` are _conditioning variables_ and the `y` and `x` are
called the primary variables. At least one primary variable must be
specified. (Unlike modelling functions, `*` and `+` in conditioning
variables are treated identically in lattice. Thus, these are valid
trellis functions in lattice.)

The formula can include method calls like `sqrt()` or `log()` to
transform individual variables.

```{r, eval = FALSE}
~ x
log(z) ~ x * y | a * b * c
```

Some examples.

|      Formula            |  Lattice Plot                            |
|------------------------:|-----------------------------------------:|
| `z ~ x * y`             | Three dimensional plot                   |
| `y1 + y2 ~ x`           | Plot two $y$'s versus $x$                |
| `y ~ x1 + x2`           | Plot $y$ versus two $x$s                 |
| `y ~ x| g1 + g2 + ...`  | Plot $y$ versus $x$ conditioned on $g$s  |
| `y ~ x| g1 * g2 * ...`  | Plot $y$ versus $x$ conditioned on $g$s  |


Mostly, the $g$'s are factors or _shingles_.

Shingles are a generalization of factors for continuous
variables. They are a way of discretizing continuous values (using
`cut`, for example), except that the intervals may overlap.  So an
observation may be assigned to more than one interval.

The formula `y ~ x | g` indicates that several plots should
be generated, showing $y$ against $x$ for each level of $g$.

Each unique combination of the levels of the conditioning variables
determines a _packet_, a subset of the primary variables for
  that combination.

___

## 10.2.3. Lattice Equivalents of Base Plots


|  Lattice Func.  |  Lattice Description             | Base Analog  |
|:----------------|:---------------------------------|-------------:|
| `barchart`      | Barcharts                        | `barplot`    |
| `bwplot`        | Boxplots                         | `boxplot`    |
| `densityplot`   | Conditional kernel density plots |              |
| `dotplot`       | Dotplots                         | `dotchart`   |
| `histogram`     | Histograms                       | `hist`       |
| `qqmath`        | Quantile–quantile plots          | `qqnorm`     |
| `stripplot`     | Strip plots                      | `stripchart` |
| `qq`            | Quantile–quantile plots          | `qqplot`     |
| `xyplot`        | Scatterplots                     | `plot`       |
| `levelplot`     | Level plots                      | `image`      |
| `contourplot`   | Contour plots                    | `contour`    |
| `cloud`         | 3-dimensional scatterplot        |              |
| `wireframe`     | 3-dimensional surfaces           | `persp`      |
| `splom`         | Scatterplot matrices             | `pairs`      |
| `parallel`      | Parallel coordinate plots        |              |


Various types of plots can be obtained by means of plot type
specifications (argument `type`).


| Type         |  Effect                      | Panel function  |
|:-------------|:-----------------------------|----------------:|
| `"p"`        | Plot points                  |                 |
| `"l"`        | Join points by line          |                 |
| `"b"`        | Both lines and points        |                 |
| `"o"`        | Points and lines overlaid    |                 |
| `"S"` `"s"`  | Plot as step function        |                 |
| `"h"`        | Drop lines to origin         |                 |
| `"a"`        | Join lines after averaging   | `panel.average` |
| `"r"`        | Join regression line         | `panel.lmline`  |
| `"smooth"`   | Plot LOESS smooth            | `panel.loess`   |
| `"g"`        | Plot a reference grid        | `panel.grid`    |

___

## 10.2.4. Annotation

Extra info can be added to a plot by specifying a _panel function_ via
the `panel` arguments.

For example, the default panel function for an `xyplot` is
`panel.xyplot()`.

This draws the default contents for a panel, so you want to call this
as part of a custom panel function"

```{r, eval = FALSE}
xyplot(mpg ~ wt, data = mtcars, panel=function(...)
       panel.xyplot(pch = 10, ...))
```

The arguments differ from function to function (see docs). Common
ones: `x`, `y` and `z` and `subscripts` (indices to subset for each
panel).

Function hooks, such as a _prepanel function_ hook to control the size
  and scaling of panels (maybe make space for some text you want to
  show) and a _strip function_ for controlling what gets drawn in
  the strips of a lattice plot.

___

## 10.2.5. A Quick Tour.

```{r}
library(lattice)
data(iris)
```

Creating a separate panel for each plot

```{r}
histogram(~Sepal.Length + Sepal.Width +
          Petal.Length + Petal.Width | Species,
          data=iris)
```

Plot all three species in a single panel.

```{r}
densityplot(~Sepal.Length + Sepal.Width +
            Petal.Length + Petal.Width, group=Species,
            data=iris)
```

Plot without the points.

```{r}
densityplot(~Sepal.Length + Sepal.Width +
            Petal.Length + Petal.Width, group=Species,
            data=iris, plot.points=FALSE, ref=TRUE)
```

Add in a legend

```{r}
densityplot(~Sepal.Length + Sepal.Width + Petal.Length +
            Petal.Width, group=Species, data=iris,
            ref=TRUE, plot.points=FALSE,
            auto.key=list(columns=3))
```

Some formula details: `*` and `+` are treated similarly only in the
conditioning part of the formula!

```{r}
densityplot(~Sepal.Length * Sepal.Width * Petal.Length *
            Petal.Width, group=Species, data=iris,
            ref=TRUE, plot.points=FALSE,
            auto.key=list(columns=3))
```

An example normal quantile plot:

```{r}
qqdata <- data.frame(x = c(rnorm(100), qnorm(ppoints(100))),
                     y = c(rep("x", 100),
                     rep("Normal Quantile", 100)))
qq(y ~ x, data=qqdata, col="red", lwd=3)
```
___

## 10.2.6. Famous Barley data

See for example `?barley`. This is a data set containing total yield
in bushels per acre for 10 varieties at 6 sites in each of two years.

```{r}
str(barley)
```

We can examine some boxplots.

```{r}
bwplot(yield ~ variety, data = barley)
```

```{r}
bwplot(variety ~ yield, data = barley)
```

```{r}
bwplot(variety ~ yield | year, data = barley)
```

Some dot plots.

```{r}
dotplot(variety ~ yield, data = barley)
```

We add groups distinguished by color

```{r}
dotplot(variety ~ yield, data = barley, groups = year,
        col=c("red", "blue"))
```

Condition on site.

```{r}
dotplot(variety ~ yield | site, data = barley)
```

Group by year and then condition on site.

```{r}
dotplot(variety ~ yield | site, data = barley, groups = year)
```

We need to add color to see difference.

```{r}
dotplot(variety ~ yield | site, data = barley, groups = year,
        col=c("red", "blue"))
```

Note that conditioning on both makes things less obvious.

```{r}
dotplot(variety ~ yield | site * year, data = barley)
```

So go back to condition on site and group by year and add a legend

```{r}
dotplot(variety ~ yield | site, data = barley, groups = year,
        col=c("red", "blue"),
        auto.key=list(columns=2, col=c("red", "blue")))
```


The plot symbol color is screwed up in the previous dotplot. So we
better have a separate symbol for each.

```{r}
dotplot(variety ~ yield | site, data = barley, groups = year,
        pch=c(2, 4),
        col=c("red", "blue"),
        auto.key=list(columns=2, col=c("red", "blue")))
```

Hmmm. Plot symbol still screwed up, so build legend manually

```{r}
dotplot(variety ~ yield | site, data = barley, groups = year,
        col=c("red", "blue"),
        pch=c(2, 4),
        key=list(text=list(c("1932", "1931")),
                 columns=2,
                 col=c("red", "blue")))
```


We can save plots to various formats such as pdf, png etc.

```{r}
pdf(file="yield.pdf", paper="letter", width=8, height=10.5)
dotplot(variety ~ yield | site, data = barley, groups = year,
        col=c("red", "blue"),
        pch=c(2, 4),
        key=list(text=list(c("1932", "1931")),
                 columns=2,
                 col=c("red", "blue")))
dev.off()
```
___

## 10.2.7. Trellis Plots

One nice thing about `lattice` (and `ggplot`) is that the plots are
actually objects.  They don't plot until you `print` them.


```{r}
## Incremental updating
tp1 <- densityplot(~Sepal.Length+Sepal.Width+
  Petal.Length+Petal.Width,group=Species,data=iris)
print(tp1)
```

You can also update the object (will not affect `tp1` above.)

```{r}
tp2 <- update(tp1, plot.points=FALSE, ref=TRUE)
plot(tp2)
```

Add the legend.

```{r}
tp3 <- update(tp2, auto.key=list(columns=3))
plot(tp3)
```

You can examine `tp3` like any other object.

```{r}
names(tp3)
tp3$legend
```

And even pick off one of the plots.

```{r}
tp3[1]
```
___

## 10.2.8. X-Y plots

```{r}
tp = xyplot(lat~long, data=quakes, pch=".")
update(tp,
  main="Earthquakes in the Pacific Ocean\n(since 1964)")
```

We can apply a 3-level shingle, no overlap.

```{r}
depthgroup <- equal.count(quakes$depth, number=3,
                          overlap=0)
magnitude <- equal.count(quakes$mag, number=2,
                         overlap=0)
xyplot(lat ~ long | depthgroup*magnitude,
       data=quakes,main="Fiji Earthquakes",
       ylab="latitude", xlab="longitude",
       pch=".",
       scales=list(x=list(alternating=c(1,1,1))),
       between=list(y=1),
       par.strip.text=list(cex=0.7),
       par.settings=list(axis.text=list(cex=0.7)))
```

___

## 10.2.9. Lattice Summary

- Improves on traditional graphics by making optimum use of display
    area by providing very fine control over layouts and taking care
    of details for you
- Better looking graphics (relatively speaking!)
- Powerful extensions are possible
- The output can be annotated, edited, saved and computed with. This
  is key, because it helps you build up plots.
- Originally, a replacement for `Trellis` graphics in Splus to produce
  a whole series of plots.

___

## Session Info

```{r}
sessionInfo()
```






