---
title: 17.1. S3 Methods
output: slidy_presentation
---

Towards the end of last class, we talked about S3 classes. S3 classes
pervade much of `stats` package, included in base R. In particular, S3
methods are used for printing, plotting and predicting on model
objects. 

## 17.1.1 Defining S3 methods

When you have an S3 class, generics like `print` and `plot` are
already available. So you don't need to do anything special if you
want to invoke a `print` or a `plot` method for your object.

For example, consider a data frame that can be plotted:

```{r}
d <- data.frame(a = 1:5, b = (1:5)^2)
class(d)
```
If we invoke plot on it, it works fine.

```{r, error = TRUE}
plot(d)
```

___

As we discussed in a previous lecture, what happened behind the scenes
is that `graphics:::plot.data.frame` was called as can be seen from:

```{r}
methods(plot)
```

and

```{r}
getAnywhere("plot.data.frame")
```
___

Now suppose we want to use `ggplot2` instead of base plot for the
scatterplot.

One way we could do this is to change the class.

```{r}
class(d) <- c("ggplot_able", class(d))
class(d)
```

Now we have to define our own method for plotting, borrowing from
`graphics:::plot.data.frame`.

```{r}
plot.ggplot_able <- function(x, ...) {
    stopifnot(require("ggplot2"))
    plot2 <- function(x, xlab = names(x)[1L], ylab = names(x)[2L],
        ...) qplot(x[[1L]], x[[2L]], xlab = xlab, ylab = ylab,
        ...)
    if (!is.data.frame(x))
        stop("'plot.ggplot_able' applied to non data frame")
    if (ncol(x) == 1) {
        qplot(x[[1L]], ...)
    } else if (ncol(x) == 2) {
        plot2(x, ...)
    } else {
        stop("plot.ggplot_able: Don't know what to do")
    }
}
```

And try it.

```{r}
plot(d)
```

It works.

But what if you have a new method that _you_ want to write? For
example, a function for returning a particular row in a data frame.

Use `UseMethod`.

```{r}
nthRow.data.frame <- function(x, n) {
    if (0L < n && n <= nrow(x))
        x[n, ]
    else
        stop("nthRow.data.frame: Bad value for n")
}

nthRow <- function(x, n) UseMethod("nthRow")
```

And we can check.

```{r}
nthRow(d, 2)
```


## 17.1.2. Ramifications

1. Because of this association of the period with methods for S3
   objects, it is best to avoid periods in names of functions, even
   though it is legitimate; it makes those functions look like methods
   for S3 objects. So many people have started using camel-case or
   underscores to avoid such confusion.

2. There can be issues if two different packages define the same S3
   classes and methods for them. A recent example was the case of
   `mgcv` and `gam` both of which defined S3 classes called `gam` and
   methods. 
   
3. The S3 system provides all sorts of opportunities for you to shoot
   yourself in the foot. Consider for example the following.
```{r, error = TRUE}
class(d) <- "foobar"
plot(d)
```
What happened?  We changed the class attribute of `d` to be something
other than a `data.frame` and so plot went ahead and called a default
method `plot.default`, the catch-all. And all hell broke loose.

Since class is merely an attribute in the `S3` system. nothing stops
you from changing it. So there are no guarantees with this type of
class system.

A more formal system called `S4` overcomes these issues.


## 17.1.3. S3 in packages

When you define new methods for S3 classes in packages, you need to
make R aware of these via Roxygen tags.  Let's take an example. 

Consider the `plot` method defined for `glmnet` (file
`R/plot.glmnet.R`) and `cv.glmnet` (file `R/plot.cv.glmnet.R`) objects
in the package `glmnet`.

___

The file `plot.glmnet.R` contains the following in the Roxygen section.

```{r, eval = FALSE}
#' @examples
#' x=matrix(rnorm(100*20),100,20)
#' y=rnorm(100)
#' g2=sample(1:2,100,replace=TRUE)
#' g4=sample(1:4,100,replace=TRUE)
#' fit1=glmnet(x,y)
#' plot(fit1)
#' plot(fit1,xvar="lambda",label=TRUE)
#' fit3=glmnet(x,g4,family="multinomial")
#' plot(fit3,pch=19)
#' @method plot glmnet
#' @export
plot.glmnet=function(x, xvar=c("norm","lambda","dev"),label=FALSE,...){
  xvar=match.arg(xvar)
 plotCoef(x$beta,lambda=x$lambda,df=x$df,dev=x$dev.ratio,label=label,xvar=xvar,...)
}
```

Similarly, the file `plot.cv.glmnet.R` contains:

```{r, eval = FALSE}
#'
#' @method plot cv.glmnet
#' @export
plot.cv.glmnet=function(x,sign.lambda=1,...) {
    Stuff_deleted...
}

```

Note how this also allows one to clearly designate what is the class
name and what is the method name: the above clearly states that we are
defining a method `plot` for objects of class `cv.glmnet`. 

So if our `nthRow` method were part of a package, we would have the
following Roxygen documentation

```{r, eval = FALSE}
#' @method nthRow data.frame
```

Once you have marked up the Roxygen documentation as shown, running
`devtools::document()` for the package will generate an entry in the
`NAMESPACE` file such as:

```{r, eval = FALSE}
S3method(plot,glmnet)
S3method(plot,cv.glmnet)
```

## 17.1.4. Session Info

```{r}
sessionInfo()
```

