---
title: "9.2. R Packages"
output:
     slidy_presentation
---
```{r requirements, echo = FALSE, message = FALSE}
## Install a package if not already installed
installIfNeeded <- function(packages, ...) {
    toInstall <- packages[!(packages %in% installed.packages()[, 1])]
    if (length(toInstall) > 0) {
        install.packages(toInstall, repos = "https://cloud.r-project.org")
    }
}

## Ensure packages exist and activate them
needPackages <- function(packages) {
    installIfNeeded(packages)
    for (x in packages) {
        library(x, character.only = TRUE)
    }
}
needPackages(c("tidyverse", "pryr"))
```


By now you know that R provides many useful functions for creating,
building, manipulating and learning about packages.  

Packages can be installed from CRAN via via `install.packages()`
function. See help on that function. (Bioconductor has its own way of
installing packages using a different function.) 

Later, we'll see how you can install packages you develop (before it
goes on CRAN) for testing etc.

A quick tour.

## 9.2.1. What packages are installed?

```{r}
x <- installed.packages()
str(x)
colnames(x)
head(rownames(x))
```

## 9.2.2. R Repositories

By default, you will be given a chance to set a repository or one will
be set for you. The site [CRAN](https://cloud.r-project.org) shows all
the repos you can use. The cloud repository makes most sense!

You can set respository using `setRepositories()` but a more common
way is to set an option:

```{r}
options(repos = "https://cloud.r-project.org")
```

will always work.


## 9.2.3. What packages are available on CRAN at this instant?

```{r}
y <- available.packages()
dim(y)
```

```{r}
head(y, 2)
tail(y, 2)
```

Note that you can compute on any of this information. In `ass2.Rmd`,
there is a code chunk as follows.

## 9.2.4. What packages are installed on my machine?

```{r}
y <- installed.packages()
dim(y)
```

## 9.2.5. Install/Download/Remove Packages

R provides `install.packages()`, `download.packages()` and
`remove.packages()` to perform these functions.

## 9.2.6. Installing needed packages on demand

Note that you can compute on any of this information. In the markdowns
for this class, there is a code chunk as below. 

```{r, eval = FALSE}
## Install a package if not already installed
installIfNeeded <- function(packages, ...) {
    toInstall <- packages[!(packages %in% installed.packages()[, 1])]
    if (length(toInstall) > 0) {
        install.packages(toInstall, repos = "https://cloud.r-project.org")
    }
}

## Ensure packages exist and activate them
needPackages <- function(packages) {
    installIfNeeded(packages)
    for (x in packages) {
        library(x, character.only = TRUE)
    }
}
needPackages(c("tidyverse", "pryr"))
```

What is it doing? It ensures that the libraries `tidyverse` is
installed before proceeding. Furthermore, it will only install
`tidyverse` if it is not already installed.

Note the use of the `...` argument that is used to pass on the `repos`
argument!

## 9.2.7. Where does my R look for packages?

Where actually are the installed packages on my machine?

```{r}
.libPaths()
.Library
.Library.site
```

From the help:

> `.Library` is a character string giving the location of the
> default library, the `library` subdirectory of `R_HOME`.
>
> `.Library.site` is a (possibly empty) character vector giving the
> locations of the site libraries, by default the `site-library`
> subdirectory of `R_HOME` (which may not exist).
>
> `.libPaths` is used for getting or setting the library trees that
> R knows about (and hence uses when looking for packages).

## 9.2.8. Add my own package tree.

One can manipulate library paths during development, for testing,
before installing in a standard place.

For example, I may want to install my packages in my home directory
under a directory `RLIBS` before I actually install it for all users
on my computer.

```{r}
.libPaths("~/RLIBS")
.libPaths()
```

On windows and Unix machines, when you install packages, you are given
a choice to add a local package tree in your home directory. This is
useful because you may not have permission to install in the system
areas.

## 13.2.2.9. Information about specific packages

```{r}
packageDescription("ggplot2")
packageVersion("ggplot2")
```

## 9.2.10. Learning to use a package

To use a package interactively, one uses `library()` or
  `require()`. Both will modify the search path, but `require` will
  return a logical value indicating if the loading was successful.

Sometimes, you want to know all the exported and documented objects/functions
in a package.

```{r}
library(help = "tibble")
```
works, but is not always convenient. Often, one has to examine the
help page on each function for details.

The above will also display messages on _vignettes_ etc. Vignettes are
the modern way to introduce the package to a user. Highly
recommended.

```{r, eval = FALSE}
vignette("tibble", package = "tibble")
```

You don't need to attach the package to look at a vignette.

## 9.2.11. Which package exports a specific function?

For example, which package provides `aggregate`?

```{r}
packageName(environment(aggregate))
```

Of course, in the above I am assuming that you have not defined your
own `aggregate` function!

The function `where` in the package `pryr` provides this info in a
more useful format.

```{r}
installIfNeeded("pryr")
library(pryr)
where("aggregate")
```

## 9.2.12. Session Info
```{r}
sessionInfo()
```
