---
title: "2.1. R Objects"
output:
  slidy_presentation
---

For data analysis, one needs to handle data that contains homogeneous
or heterogeneous variables. Data may be also 1-dimensional or
multidimensional.

The following table (from _Advanced R_) shows the table:


|Dimension |Homogeneous      | Heterogeneous |
|----------|-----------------|---------------|
| 1-d      |Atomic vector    |List           |
| 2-d      |Matrix           |Data frame     |
| $n$-d    |Array            |               |

## 2.1.1. Types

R has a fixed set of types (`enum` in C, but shown as strings in R).
You have all the usual basic types: booleans, integers, real numbers, complex
numbers, bytes, strings. In R these are called `logical`, `integer`, `numeric` or
`double`, `complex`, `raw`, and `character` respectively.  NULL is
also used to

```{r}
1.5
typeof(1.5)
typeof(1)
typeof(1L)
typeof(1 + 0i)
typeof(quote(x)) ## symbol
typeof(quote(1 + 2)) ## language object
typeof(rnorm(10))
```

Note that there are no scalar types, the first and the last above are
both vectors!

___

All objects have a type; some types share properties, like
_vector_ness.

```{r}
is.atomic(1.5)
is.atomic(list(1,2))
typeof(list(1, "a string"))
typeof(c(1, "a string")) ## note coercion
```

The atomic types are `logical`, `integer`, `numeric`, `complex`,
`character` `raw`  and `NULL`.

## 2.1.2. Vectors

A list is a generic vector.


```{r}
is.vector(rnorm(10)) ## yes
is.vector(list(1, "a string"))  ## list is a generic vector
is.vector(c(1, "a string"))
```

There are a family of `is.x` functions where `x` is some property you
want to test.

Let us examine the result of adding some metadata to a created object
(_attributes_).

```{r}
x <- rnorm(10)
is.vector(x)
attributes(x) <- list(name = "x", value="ten normals") ## meta-data
is.vector(x) ## no, because of attributes.
```

Moral: it is possible to get fooled by function names. The only way to
be sure is to read the documentation and perhaps see its
implementation. (RStudio has a help panel where you can search for
documentation as noted in the quick introduction.)

A robust way to test for a vector is as follows.

```{r}
x <- list(x = 1, y = 2)
is.vector(x)
is.atomic(x) || is.list(x) ## correct way to test for vectorness
attr(x, "some_meta_data") <- "foobar"  ## attr sets a particular attribute!
is.vector(x)
is.atomic(x) || is.list(x) ## works correctly
```

## 2.1.3. Non-vector types

What about other types?

```{r}
f <- function(x) x + 1
typeof(f)
is.vector(f)
typeof(rnorm)
typeof(log)
rnorm
log
.GlobalEnv  ## environment
typeof(.GlobalEnv)
is.vector(.GlobalEnv)
```


## 2.1.4. Attributes, class

Meta data can be added to any object via `attributes`. It is empty for
plain objects, a named list for structures, objects from
classes. Individual attributes are extracted or set by `attr(x,
which)`.

```{r}
attributes(rnorm(10))
attributes(iris3) ## a 3-way array
attributes(iris) ## a data frame (same data, different structure)
attr(iris3, "dim")
```

Attributes can affect how some objects behave (recall `is.vector`) and
also how they are printed. From R help on `?class`:

> Types are fixed; for programming "class" is more flexible, defined at
> the R level, specifically in the `methods` package. Many R objects
> have a `class` attribute, a character vector giving the names of the
> classes from which the object _inherits_.  If the object does not have
> a class attribute, it has an implicit class, '"matrix"', '"array"' or
> the result of 'mode(x)' (except that integer vectors have implicit
> class '"integer"')...

```{r}
class(rnorm(1))
class(rnorm); typeof(rnorm)
class(log); typeof(log)
class(show); typeof(show)
```

With formally defined classes, we have slots, not attributes:
guaranteed for all objects from a class; the attributes are used to
implement slots. When we come back to objects, for OOP, we'll look at
class as programming mechanism

## 2.1.5. Examining objects

Examine objects succinctly with the very useful function `str`.

```{r}
str(iris)
str(rnorm(10))
str(list(1L, 2L))
```

While not useful for _all_ R objects, `str` is especially useful in
examining some data objects.

Typing the name of some object will also print it since R will simply
evaluate it.

```{r}
matrix
```

## Session Info
```{r}
sessionInfo()
```
