---
title: "2.4. Arrays"
output:
     slidy_presentation
---

Arrays are multi-dimensional vectors with some extra attributes. Names
can be assigned to the dimensions.  The standard R array functions
adhere to column-major order.

Two-dimensional arrays aka matrix. 

```{r}
array(1:6, dim = c(2, 3))
```

The above 2 by 3 matrix consists of three columns of 2 elements each. 

In memory it is stored in contiguous locations as `1, 2, 3, 4, 5, 6`.

Contrast with Python, where it would, by default, be stored in memory
locations as `1, 3, 5, 2, 4, 6`.
            
___

Multidimensional array. 

```{r}
a <- array(1:24, dim = c(2, 3, 4))
a
```
Here also, the elements are stored contiguously in memory, `1,
2,..., 24` and the indices $(i, j, k)$ vary as follows for those
elements, because of column major ordering.

- $i$ runs as  `r rep(1:2, 12)`
- $j$ as `r rep(c(1,1,2,2,3,3), 4)`
- $k$ as `r c(rep(1,6),rep(2,6),rep(3,6),rep(4,6))`

---

Examine attributes of an array.

```{r}
attributes(a)
```

Change the names of the dimensions.

```{r}
dimnames(a) <- list(c("a", "b"), c("A", "B", "C"), c(1, 2, 3, 4))
attributes(a)
a
```

___

One can transform a vector into array easily by setting attributes
explicitly.

```{r}
x <- 1L:24L  ## the L suffix forces integer storage
class(x)
dim(x) <- c(2, 3, 4)
x
typeof(x)
class(x)

x <- 1L:12L
class(x)
attributes(x) <- list(dim = c(3, 4))
class(x)
x
```

## 2.4.1. Subsetting
```{r}
a <- array(1:24, dim = c(2, 3, 4))
a[, , 1]
a[1, , ]
a[1, , , drop = FALSE] ##
```

So `drop` is a way to control whether the dimensions are dropped or
not. By default, it is true and therefore, the result is simplified.
But there are many cases where you'd not want that to happen,
especially with matrix operations (like matrix multiply).

```{r}
x <- matrix(1:6, nrow = 3)
print(x)
y <- x[, 2]
print(y)
class(y)
z <- x[, 2, drop = FALSE]
class(z)
print(z)
dim(z)
```

## 2.4.2. Matrices and Matrix subsetting

Matrices are, of course, very commonly used and so R has functions for
handling them, including all the standard Linear Algebra stuff.

```{r}
matrix(0, nrow = 3, ncol = 2)
matrix(1:10, nrow = 2)
matrix(1:10, nrow = 2, byrow = TRUE)
```

You can subset matrices in any appropriate way.

```{r}
timesTable <- outer(1:5, 1:5, FUN = "*")
print(timesTable)
timesTable[1:3, 1:3]
```

Matrices can also have row names and column names and they can be very
useful in writing readable code. You can also use those names in
subsetting.

```{r}
m <- matrix(1:6, nrow = 2)
rownames(m) <- paste0("r", seq_len(nrow(m)))
colnames(m) <- paste0("c", seq_len(ncol(m)))
m
m["r2", "c3"]
```

## Session Info
```{r}
sessionInfo()
```
