---
title: "8.1. Tidy Data"
output:
     slidy_presentation
---

```{r, echo = FALSE, results = 'hide', message = FALSE}
## Install a package if not already installed
installIfNeeded <- function(packages, ...) {
    toInstall <- packages[!(packages %in% installed.packages()[, 1])]
    if (length(toInstall) > 0) {
        install.packages(toInstall, repos = "https://cloud.r-project.org")
    }
}

## Ensure packages exist and activate them
needPackages <- function(packages) {
    installIfNeeded(packages)
    for (x in packages) {
        library(x, character.only = TRUE)
    }
}
needPackages(c("png", "grid", "gridExtra"))
```



```{r, echo = FALSE}
library(tidyverse)
```

Tidy data is a standard way of mapping the meaning of a dataset to
its structure. A dataset is messy or tidy depending on how rows,
columns and tables are matched up with observations, variables and
types. In tidy data:

- Each variable forms a column
- Each observation forms a row
- Each type of observational unit forms a table cell.

___

Tidy data makes it easy to do analyses although there are applications
where data may be stored in preferred formats (not necessarily tidy by
the definition above) for efficiency reasons. The reason for a lot of
data wrangling and cleaning is really to create tidy data sets because
measured variables become clear and the analysis, easy and
straightforward.

___

Consider an example tibble.

```{r}
d <- tibble(
    id = 1:4,
    gender = c("M", "F", "F", "M"),
    sbp.trt1 = c(125, 110, 100, 135),
    dbp.trt1 = c(81, 90, 80, 90),
    sbp.trt2 = c(120, 115, 114, 130),
    dbp.trt2 = c(75, 91, 82, 88)
)
d
```

Although five variables are measured for a patient: `id`, `gender`,
`sbp`, `dbp`, `treatment` (1 or 2), not all of these in a column by
themselves. For example the treatment that the patient got is along
columns. 

This is not tidy data.

Notice also how complicated it is to compute a simple thing such as
the overall average of all systolic pressure measurements.

- We have to find the two columns that relate to systolic pressure
  measurements
- We would have to sum all the values
- We would have to divide by the number of measurements. 

The natural thing we expect to do is apply a function like `mean` to a
single column and the above doesn't fit for this data.

___

```{r}
d
```

The natural format is one where have a column for the treatment
(`trt`), a column for the type of blood pressure measurement (`sbp` or
`dbp`) and a column for the blood pressure value (`bp`). If we have
that format, then we can bring the power of `dplyr` and pipes
discussed above to compute all kinds of summaries very easily. 

Tidying data also makes it easy to produce complex plots as we shall
see later. 

___


## 8.1.1. The package `tidyr`

The package `tidyr` (in `tidyverse`) provides some functions to
reshaping the data into a form you want.  The operative word here is
_reshape_!

Note: Package `tidyr` contains function `gather` which also does what
I am going to describe below, but it is now deprecated as new improved
functions have replaced it.

___

### Step 1

The `tidyr::pivot_longer` function will take columns in a multiple
columns and collapses into key-value pairs. It takes care to duplicate
all other columns in the data set. This function is useful when we
have have columns that are not variables. (That is certainly the case
for us since the columns contain treatment information: `trt1` and
`trt2`.)


```{r}
d %>%
    pivot_longer(cols = c(sbp.trt1, dbp.trt1, sbp.trt2, dbp.trt2),
                 names_to = "trt", values_to = "bp")
```

The above invocation asks `pivot_longer` to take the four blood pressure
columns and convert them into key-value pairs. The key column will be
named `trt` and the value column will be given the name `bp`.

We can see this clearly by looking at the original data and after it
has been gathered, side-by-side.

```{r, echo = FALSE}
img1 <-  rasterGrob(as.raster(readPNG("before_gather.png")), interpolate = FALSE)
img2 <-  rasterGrob(as.raster(readPNG("after_gather.png")), interpolate = FALSE)
grid.arrange(img1, img2, ncol = 2)
```

Notice how `pivot_longer` duplicated the `id` and `gender` columns to keep
the data consistent. The order in which the values appear is dictated
by the order of the arguments specified in the `pivot_longer` invocation
above.

___

Instead of telling `pivot_longer` which columns to convert to
key-value pairs, we can also tell it which columns _not to use_.

Thus, the same effect above could be achieved via:

```{r}
d %>%
    ## gather all columns but id and gender
    pivot_longer(cols = -c(id, gender), names_to = "trt", values_to = "bp")    
```

Both forms are very useful.

The `tidyverse` further provides selection facilities (via `tidyselect`
package) for specifying columns using patterns for their names, such as
  `starts_with`, `ends_with`, `one_of`, `contains`, `matches`,
  etc. See `?tidyselect::select_helpers` and examples.

Again, the same effect, via a different invocation.

```{r}
d %>%
    ## gather all columns that contain the "bp"
    pivot_longer(cols = tidyselect::contains("bp"), names_to = "trt", values_to = "bp")    
```

___ 

What happens when the columns being gathered have different types?
Let us see.

```{r, error = TRUE}
d %>%
    pivot_longer(cols = -id , names_to = "trt", values_to = "bp")   
```

This is a difference from how `tidyr::gather` behaves:

```{r}
d %>% gather(key = key, value = value, -id)
```

So in `gather`, `key` is character, and `value` is coerced, promoted
to a type that can encompass both types of value, which is character.

This is usually not a good idea, but is sometimes is a quick-and-dirty
trick for data cleaning.

___

### Step 2

So we now have a column that has isolated the treatment. However,
there is a problem. It has also combined the measurement `sbp`, `dbp`
into the treatment column. So we need to separate it out.

Enter the `tidyr::separate` function, which will separate a specified
column into several columns.

```{r}
d %>%
    pivot_longer(cols = tidyselect::contains("bp"), names_to = "trt", values_to = "bp") %>%    
    separate(col = trt, into = c("bpType", "trt"))
```

Nice, except how did `separate` know to separate `sbp.trt1` into `sbp`
and `trt1`? In other words, how did it know what to use as a
separator?

By default, `separate` uses all non-alphanumeric characters as
separators and period is one of them. In case your data set has
some other separater, that can be specified using the `sep` argument
to `separate`.

At this point, as you can verify, the data is in a tidy form. So let
us save it after making sure we `rename` the `sbp` column

```{r}
d %>%
    pivot_longer(cols = tidyselect::contains("bp"), names_to = "trt", values_to = "bp") %>%    
    separate(col = trt, into = c("bpType", "trt")) ->
    tidyD
```

___

Instead of using `separate`, we can also do the entire thing in one
fell swoop. 

```{r}
d %>%
    pivot_longer(cols = tidyselect::contains("bp"),
                 names_to = c("bpType", "trt"),
                 values_to = "bp",
                 names_pattern = "([a-z]+)\\.([a-z]+[1-2])")
```

This is often more efficient, at the slight cost of readability.

## 8.1.2. Apropos Keys and Values

The idea of organizing data using key-value pairs is fundamental and
ubiquitous in programming and data science. In fact, one can go full
monty and organize any data element this way. Databases that organize
most of this data this way, are called [Entity-Attribute-Value
(EAV)](https://en.wikipedia.org/wiki/Entity%E2%80%93attribute%E2%80%93value_model)
databases.

In such databases, the tables are very skinny, essentially three
columns:
- what is being measured (the entity)
- what attribute of the entity is being measured (the attribute)
- what value does that attribute have (the value)

The advantage is that the organization is very general, flexible,
efficient and sparse. However there are disadvantages as well.

Since one essentially has a large duffel-bag where you can shove
everything, retreiving an entity value involves a search and then
possible type conversion everytime, which can be expensive and hard to
manage. 

So the functions in `tidyr` make it easy to produce such long skinny
_tidy_ data that make summarizing operations easier.

## 8.1.3. Summaries

The tidiness makes it very easy to compute summaries etc.

1. Mean systolic blood pressure for all patients

```{r}
library(dplyr)
tidyD %>%
    filter(bpType == "sbp") %>%
    select(bp) %>%
    summarize(meanSBP = mean(bp))
```

2. Mean systolic blood pressure by treatment

```{r}
tidyD %>%
    group_by(trt) %>%
    filter(bpType == "sbp") %>%
    select(bp) %>%
    summarize(meanSBP = mean(bp))
```

3. Mean blood pressure by treatment, by gender, by type

```{r}
tidyD %>%
    group_by(trt, gender, bpType) %>%
    select(bp) %>%
    summarize(meanSBP = mean(bp))
```

## 8.1.4. Further Facilities

The package `tidyr` provides functions such as 

- `extract` which also turns each group into a new column, with
  facilities for type conversion
- `unite` which is the complement of `separate` which combines
  multiple columns into 1, etc.
  
Use `library(help=tidyr)` to see all the exported functions and also
to see vignettes.

## 8.1.5. Discussion

We saw how data transformation may be needed for analysis.

- The series of transformations of a data set in order to produce an
  analyzable version can be nicely visualized as a pipleline

- The package `dplyr` provides some nice verbs to aid in the
  transformations and analysis

- The verbs in `dplyr` may be applied uniformly to data in R or
  structures in R that are backed by remote data (i.e. data in
  databases)
  
- Tidy data makes it easy to analyze data, especially using `dplyr`
  verbs.  The package `tidyr` provides several functions to make this
  possible. A really nice large example using
  [WHO lung data](http://r4ds.had.co.nz/tidy-data.html#case-study) is
  discussed in Wickham and Gorlemund's books _R for Data Science_. We
  urge you to read it.

- Tidy data works well with `ggplot` visualization tools. We shall see
  this shortly. 

By combining all these packages, one can tackle many of the
data-related tasks and produce data that can be analyzed.

## 8.1.6. Resources

- [Case study of WHO lung
  data](http://r4ds.had.co.nz/tidy-data.html#case-study)
- Help pages for `tidyr`, `tidyselect` packages. 
- Vignette for `pivot` in package `tidyr`. See also `pivot_wider`. 

## 8.1.7. Session Info
```{r}
sessionInfo()
```
