---
title: 24.1. Performance
output:
     slidy_presentation
---

R is an extremely dynamic language. We have discussed R's scoping
rules earlier and seen how R looks up free variables dynamically. This
is because the S language (on which R is based) was developed for
exploratory analysis. 

This also means that a completely undeclared variable will not
necessarily cause an error in interactive scripts, for you might have
a variable declared and saved in a session that might be loaded before
running the script. (I would, however, say it does imply a degree of
danger and should be generally avoided in preference to explicit
saving and loading of variables.)

## 24.1.1. Constants

There is no real notion of a constant in the R language unless you
hardwire it. That is you cannot define:

```{r, eval = FALSE}
TWO <- 2L
```
and assume it is going to be treated as a constant. Your code can
actually modify `TWO` later and no questions will be asked.

Of course, if you use `2L` wherever you used `TWO`, that would indeed
be treated as a constant, but using a name for constant usually makes
code more readable!

## 24.1.2. Functions

One can change any aspect of a function: argument, body, and even
  environment. 
  
Which means that the details of a function object can only be detected
at the actual execution time!

Even looking up a variable inside a function is expensive, since one
can never know where it will come from, the current environment, or
its parent, etc. etc., or even cause an error.

Therefore, function calls are expensive, involving setting up
environments, promises, before actually executing the body. Lazy
evaluation helps but only so much. 

Which is why vectorization pays off in R programming: once the
variable is located, all one has to do is sweep over its elements
without the added overhead of a function call for each operation.

## 24.1.3. Object Class

One can change the class of objects at will _at runtime_. 

One can also convert types and there is coercion as well. 

This means techniques like _type inference_ cannot be exploited to
create efficient code. 

There are modern efforts underway to add annotations to object that
can be used to assume determined behavior, but they are not yet fully
fledged and under consideration. For example, one might specify that a
function argument or object data is of a certain type, which can let
the R byte compiler do some inference downstream.

## 24.1.4. Generic Functions

Generic functions dispatch based on type of object.  If a new class of
object is added and a generic defined for it, that means another look
up in a table for dispatch. 

Since the table itself is dynamic, it is looked up at runtime. So,
the cost can be significant.  

Let us examine this via an example. 

## 24.1.5 Timing Comparisons

Let us see this in action using a slightly modified example on [Stack
Overflow](https://stackoverflow.com/questions/16386154/is-s4-method-dispatch-slow)

```{r}
library(R6)

## Bare function
f <- function(x) NULL

## S3 method
s3 <- function(x) UseMethod("s3")
s3.integer <- function(x) NULL

## S4 class and method
A <- setClass("A", representation(a = "list"))
setGeneric("s4", function(x) standardGeneric("s4"))
setMethod(s4, "A", function(x) NULL)

## Reference class (R5) method
B <- setRefClass("B")
B$methods(r5 = function(x) NULL)

## R6 method
C <- R6Class("C",
             public = list(r6 = function(x) NULL))

## Create objects
a <- A()
b <- B$new()
r6 <- C$new()

library(microbenchmark)
knitr::kable(
           summary(
               microbenchmark(
                   bare = NULL,
                   fun = f(),
                   s3 = s3(1L),
                   s4 = s4(a),
                   r5 = b$r5(),
                   r6 = r6$r6(),
                   times = 1000
               )))
```

## 24.1.6. Projects for Speeding up R

There are several projects that are dedicated to improving R
performance.

___

###  Pretty Quick R

[Pretty Quick R](http://www.pqr-project.org/) by Radford Neal aims
to speed up R. In particular, per the project statement:

- Multiple processor cores can automatically be used to perform some
  numerical computations in parallel with other numerical
  computations, and with the thread performing interpretive
  operations. No changes to R code are required to take advantage of
  such computation in “helper threads”.

- pqR makes a better attempt at avoiding unnecessary copying of
  objects, by maintaining a real count of “name” references, that can
  decrease when the object bound to a name changes. Further
  improvements in this scheme are expected in future versions of pqR.

- Some operations are avoided completely in pqR — for example, in pqR,
the statement for (i in 1:10000000) ... does not actually create a
vector of 10000000 integers, but simply sets i to each of these
integers in turn.

The problem is that this version is always behind the latest R
version. Current pqR is based on R version 2.15.x.

Several of these improvements have made into the usual R itself
now. And recent releases of R use `altrep` (alternative internal
representation of objects) which yield improvements in both size of
objects and speed (representation of sequences, run length encoded
vectors, sortedness, `NA`s present, etc.)

As an example, consider the following:

```{r, eval = FALSE}
## Loop over an index 1 million times
for (i in 1:1e6) {
    ...
}
```
Previous R would actually allocate a million-element vector, but now
it just uses a sequence with a start and end and executes the loop.

Similarly, if meta-data on whether `x` is sorted causes a clause like
`x > 10` to return much faster.

___

### Renjin

[Renjin](http://www.renjin.org) is a Java based interpreter for the R
language.

- Better support for big data with facilities to interact with data
  stored in the cloud
- Uses deferred computation and just-in-time compilation to achieve
better performance

Not all CRAN packages are certified to work with Renjin. So that is
something to watch for.

___

### FastR

FastR is an implementation of the R Language in Java atop Truffle, a
framework for building self-optimizing AST interpreters.

[FastR](https://github.com/graalvm/fastr) is an implementation of the
R Language in Java atop Truffle, a framework for building
self-optimizing AST interpreters.

FastR is intended eventually to be a drop-in replacement for GNU
R. Currently, however, the implementation is incomplete. Notable
limitations are:

- Graphics support is mostly missing, limited to output to the "pdf" device.

- Many packages either do not install, particularly those containing
  native (C/C++) code, or fail tests due to bugs and limitations in
  FastR. In particular popular packages such as data.table and Rcpp
  currently do not work with FastR.

___

### Academic Research

There is also work afoot on where there are possibilities for
improvement, particularly in the

- [DSC](https://www.r-project.org/conferences.html)
- [Radford Neal](http://www.cs.toronto.edu/~radford/)
- [Jan Vitek](http://janvitek.org/)

And companies such as Oracle and Microsoft also have their own teams
working on R.

Chapters 23 and 24 of [Advanced R](https://adv-r.hadley.nz) also have
useful details on performance.
