---
title: 'Stat 290: Assignment 1: Q4'
output:
  html_document:
    df_print: paged
---

## Q4 (Huffman encoding) [9 pts]

  A classic programming example is Huffman encoding. See the famous
  book by [Abelson, Sussman, and
  Sussman](https://mitpress.mit.edu/sites/default/files/sicp/full-text/book/book-Z-H-16.html#%_sec_2.3.4). That
  example is in Scheme but we build an R version. The main difference
  between a purely functional implementation and the one in R is the
  avoidance of recursion, particularly because R does not support
  things like tail recursion. Also, we make do with generic lists in
  R.

  For this exercise, complete parts of the code below; you will need the file
  `totf.txt` (included on coursework). For _your_
  (not our) checking you need to make sure the file is available in
  the same directory as this markdown.

```{r q4_setup, eval = TRUE}
##
## Make a leaf node (also a codeTree) with given symbol and given weight
## Just a list representation.
makeLeaf <- function(symbol, weight)
    list(list(symbol=symbol, weight=weight))

## Leaves have no left or right sub-trees
is.leaf <- function(object) is.null(object[[1]]$left) && is.null(object[[1]]$right)

## Accessors for getting at symbol and weight
symbol <- function(object) object[[1]]$symbol
weight <- function(object) object[[1]]$weight

##
## Construct a codeTree with given left and right code trees
##
makeCodeTree <- function(left, right)
    list(list(left=left,
              right=right,
              symbol=c(symbol(left), symbol(right)),
              weight=weight(left) + weight(right)))

## Left and right branch accessors
leftBranch <- function(tree) tree[[1]]$left
rightBranch <- function(tree) tree[[1]]$right

##
## Traverse the tree per bit
##
chooseBranch <- function(bit, branch) {
    if (bit == 0)
        leftBranch(branch)
    else if (bit == 1)
        rightBranch(branch)
    else
        stop(paste("bad bit in chooseBranch", bit))
}
```

__4a__. Write code to create a vector of ordered frequencies from the
    character vector \texttt{helloWorld}. (1 liner)

```{r q_4a, eval = TRUE}
helloWorld <- c("H", "e", "l", "l", "o", " ",
                "W", "o", "r", "l", "d", "!")

helloFreq <- sort(table(helloWorld))
```

__4b__. Create a list of Leaf nodes each of which represents a character
    of the `helloWorld` vector. _Hint: `lapply`_. If you use `mapply`
    be sure to read the documentation

```{r q_4b, eval = TRUE}
helloLeafList <- lapply(seq(helloFreq),
                        function(i) makeLeaf(weight=helloFreq[i],
                                             symbol=names(helloFreq)[i]))
```

or 

```{r q_4b2, eval = TRUE}
helloLeafList <- mapply(makeLeaf, names(helloFreq), helloFreq, SIMPLIFY = FALSE)
```
__4c__. Complete the function `encode` and `decode` (one line each).

```{r q_4c_encode, eval = TRUE}
##
## Given a code, encode the vector of chars
## and return a vector of integers representing bits
##
encode <- function(codeTree, text) {
    result <- list()
    n <- length(result)
    for (ch in text) {
        subTree <- codeTree
        while (!is.leaf(subTree)) {
            n <- n + 1
            if (ch %in% symbol(leftBranch(subTree))) {
                result[[n]] <- 0
                subTree <- leftBranch(subTree)
            } else {
                result[[n]] <- 1
                subTree <- rightBranch(subTree)
            }
        }
    }
    ## 4c. return a vector of integers
    unlist(result)
    ## End of 4c.
}
```

```{r q_4c_decode, eval = TRUE}
##
## Given a codeTree, decode the vector of bits
## and return a character vector
##
decode <- function(bits, codeTree) {
    result <- list()
    n <- length(result)
    currentBranch <- codeTree
    for (bit in bits) {
        currentBranch <- chooseBranch(bit, currentBranch)
        if (is.leaf(currentBranch)) {
            n <- n + 1
            result[[n]] <- symbol(currentBranch)
            currentBranch <- codeTree
        }
    }
    ## 4c. return a vector of characters
    unlist(result)
    ## End of 4c.
}
```

__4d__. Complete the function `createCodeTree` (no more than 4 or 5 lines).

```{r q_4d, eval = TRUE}
##
## Given a list of codeTrees ordered by weight, combine the two lowest
## weight codeTrees and insert it in the rest of the list, preserving
## the weight orderings
##
combine <- function(trees) {
    insert <- function(item, target) {
        if (length(target) == 0) {
            item
        } else {
            x <- local({tmp <- sapply(target, weight); split(seq(target), tmp > weight(item))})
            c(target[x[["FALSE"]]], list(item), target[x[["TRUE"]]])
        }
    }
    if (length(trees) < 2)
        trees
    else {
        ## Combine the first two
        combined <- makeCodeTree(trees[[1]], trees[[2]])
        insert(combined, trees[-(1:2)])
    }
}

##
## Create a single codeTree from a list of codeTrees
## Typically arguments is a list of Leaf nodes
## Hint: reduce to a singleton using combine
##

createCodeTree <- function(trees) {
    ## 4d. Complete the function....
    ## return a singleton tree
    while(length(trees) != 1) {
        trees <- combine(trees)
    }
    trees
    ## end of 4d.
}
```

__4e__. Create a code tree for the `helloWorld` vector.

```{r q_4e, eval = TRUE}
codeTree <- createCodeTree(helloLeafList)
```

__4f__. Encode and decode `helloWorld` and check that the
    encoding and decoding process are consistent (3 lines max).

```{r q_4f, eval = TRUE}
encoded <- encode(codeTree, helloWorld)
decoded <- decode(encoded, codeTree)
identical(helloWorld, decoded)
```

__4g__. Decode the message in `testMessage` and print it as _one_
string using the `cat` function.

```{r q_4g_setup, eval = TRUE}
testMessage <- scan(
    textConnection(
        "0 0 0 1 0 1 0 1 1 0 1 0 1 1 1 1 1 0 1 1 1 0 1 1 1 0 0 1 1 1 1 1 0 1 1 0 1
        0 1 1 1 1 1 0 1 1 0 1 1 0 1 0 0 1 1 0 0 1 1 0 0 0 1 0 0 1 0 0 0 1 1 1 1 1
        0 0 0 1 1 1 1 1 1 1 0 1 1 1 0 0 0 0 1 0 0 0 1 0 0 0 0 1 0 0 1 1 1 0 1 1 0
        0 1 0 1 1 0 1 1 0 1 1 0 1 1 1 1 1 0 0 0 0 0 1 1 0 1 0 1 0 0 1 1 0 1 0 1 0
        1 1 0 0 1 1 0 1 1 0 1 1 1 1 0 0 1 0 1 0 1 1 1 1 0 1 1 1 1 0 1 1 0 0 0 1 0
        0 1 0 0 0 0 0 1 1 1 0 0 0 1 0 0 1 1 1 0 1 1 1 1 0 0 1 1 1 1 0 1 1 1 1 0 0
        1 0 1 0 0 1 1 1 1 0 1 0 1 1 0 0 1 1 1 0 0 0 0 1 1 1 1 0 1 0 1 1 1 0 0 1 0
        0 0 1 1 0 1 0 1 0 1 0 1 0 1 1 0 0 1 1 0 1 0 1 0 0 0 0 1 0 0 1 1 1 1 1 0 0
        1 0 1 1 1 0 0 0 0 1 1 0 0 1 1 1 1 0 1 1 0 0 1 0 1 1 0 1 1 1 1 1 0 0 0 0 0
        1 1 0 1 0 1 0 0 1 1 0 1 0 1 0 1 1 0 0 1 1 0 1 1 0 1 1 1 1 0 0 1 0 1 0 1 1
        1 1 0 1 1 1 1 0 0 1 1 1 1 0 1 1 0 0 1 0 1 1 0 1 1 1 1 1 1 0 1 0 1 1 0 0 1
        1 1 0 0 0 0 1 1 1 1 0 1 0 1 1 1 0 0 1 0 0 0 1 1 0 1 0 1 0 1 0 1 0 1 1 0 0
        1 1 0 1 0 1 0 0 0 0 1 0 0 0 1 0 1 0 1 1 0 0 0 0 0 0 1 0 1 0 1 1 0 1 0 1 1
        1 1 1 0 1 1 1 0 1 0 0 0 0 0 1 1 1 0 0 1 1 1 1 1 0 1 1 0 0 1 0 0 0 0 1 1 1
        0 0 0 0 1 1 0 0 0 0 0 1 1 0 1 0 0 1 1 1 1 0 1 0 0 0 0 1 1 0 0 1 1 1 0 1 1
        1 0 1 1 0 0 1 1 0 1 0 0 0 0 1 1 0 0 1 1 1 0 0 0 1 0 0 0 1 0 1 1 1 1 0 0 1
        1 1 1 0 0 0 1 1 0 0 1 1 0 0 0 0 0 0 0 1 0 1 0 0 1 1 1 0 0 1 1 0 1 0 0 0 1
        1 1 1 1 0 1 0 1 0 0 1 1 1 0 0 1 1 1 0 1 1 1 0 0 0 0 0 1 1 1 1 0 1 0 1 1 0
        0 1 1 1 0 0 0 0 1 1 0 1 1 0 1 0 1 1 1 1 1 0 1 1 1 1 0 0 1 1 1 1 0 1 0 1 1
        0 1 1 1 0 1 0 0 0 1 0 1 1 1 1 0 1 0 1 0 1 1 1 1 1 1 1 1 0 1 1 1 0 0 1 1 1
        1 0 0 1 1 1 1 0 1 0 1 1 0 1 0 0 0 0 1 0 0 0 1 1 0 0 0 1 1 1 1 1 1 0 0 1 1
        1 0 1 0 1 0 1 0 0 1 0 1 1 0 1 1 0 1 0 0 1 0 0 0 1 1 1 0 1 1 1 0 0 0 0 0 1
        1 1 0 0 0 1 0 0 1 0 0 1 0 1 1 0 1 0 0 0 1 0 1 0 1 1 0 0 0 0 0 0 1 0 1 1 0
        1 1 1 1 1 0 0 0 0 1 1 0 0 1 1 0 0 1 1 1 1 1 0 1 1 1 1 0 0 1 0 0 1 0 0 0 0
        0 0 1 1 1 1 0 1 1 0 1 0 1 1 0 1 1 0 1 0 1 1 1 1 0 1 1 1 1 1 1 1 0 1 1 0 0
        1 1 0 1 1 0 0 0 0 1 1 0 0 0 0 0 1 1 0 1 0 0 0 1 0 1 1 1 1 1 1 0 1 1 1 0 0
        1 1 1 1 1 0 1 1 1 1 0 1 0 0 0 0 1 1 1 0 0 0 0 1 1 0 0 0 0 0 1 1 0 1 0 0 1
        0 1 0 0 1 1 0 1 1 1 0 0 1 0 0 0 0 1 0 1 1 0 0 0"),
    what=integer(0))

testMessageFreq <- scan(
    textConnection(
        " 1  1  1  1  1  2  2  2  2  2  2  3  3  3  4  4  4  4  5  5  6  8  9 10 13 13 
         14 14 15 18 22 41"),
    what = integer(0))

names(testMessageFreq) <- c("A", "E", "v", "w", "y", "-", ",", "b", "h", "I",
                            "P", "\n", ".", "1", "0", "c", "d", "p", "f", "g",
                            "l", "m", "u", "i", "e", "s", "o", "r", "n", "t",
                            "a", " ")
```

```{r q_4g, eval = TRUE}
leafList <- lapply(seq(testMessageFreq),
                   function(i) makeLeaf(weight=testMessageFreq[i],
                                        symbol=names(testMessageFreq)[i]))
codeTree <- createCodeTree(leafList)
decoded <- decode(testMessage, codeTree)

cat(paste(decoded, collapse=""))
```

__4h__. Explain what happens in the lines
```{r, eval = FALSE}
x <- local({tmp <- sapply(target, weight); split(seq(target), tmp > weight(item))})
c(target[x[["FALSE"]]], list(item), target[x[["TRUE"]]])
```
of the function `combine` in part d. Particularly state what happens for the extreme cases.

```{r q_4h, eval = FALSE}
## We split the codetrees in tree into two groups,
## ones with weight <= that of item, and those above it.
## The result is a list with *named* elements "FALSE" or "TRUE"
## respectively. In the extreme cases, the list will have one
## element rather than two, all above or all below.
## However, by using the names to access the elements,
## we get an empty list if no elements exist.
## So the code won't fail.
```

__4i__. Encode and decode `totfHex` and check they encode and decode
consistently.  Ignoring the storage of the code tree and other
metadata, what is the approximate compression (compressed size /
original size) you get for the _A Tale of Two Cities_ snippet? (Assume
8 bits/character). Print this ratio.

```{r q_4i_setup, eval = TRUE}
totf <- readBin(con="totf.txt", what=raw(0), n=10^6)
totfHex <- as.character(totf)
```

```{r q_4i, eval = TRUE}
totfFreq <- sort(table(totfHex))
leafList <- lapply(seq(totfFreq),
                   function(i) makeLeaf(weight=totfFreq[i],
                                        symbol=names(totfFreq)[i]))
codeTree <- createCodeTree(leafList)
encoded <- encode(codeTree, totfHex)
decoded <- decode(encoded, codeTree)
stopifnot(identical(totfHex, decoded))

print(length(encoded)/8/length(totfHex))
```

## Session Info

Leave everything below as is.

```{r}
sessionInfo()
```
