---
title: 'Stat 290: Assignment 1: Q1'
output:
  html_document:
    df_print: paged
---

## Q1 (Short answers) [11 pts]

No more than 2 sentences.

In a fresh interactive R session...

__1a__. What does `x = 2` do?

```{r q_1a, eval = FALSE}
##Assigns the value '2' to the variable 'x' as Type:double for default. The associativity is right-to-left.
```

__1b__. What does `(x = 5)` do?

```{r q_1b, eval = FALSE}
##Assigns the variable 'x' to value '5'. The associativity is left-to-right.
```

__1c__. What does `(x == 2)` do?

```{r q_1c, eval = FALSE}
##This is a comparison operator (left-to-right associativity) that checks whether the variable 'x' is equal to the double '2'. This returns a boolean value of T/F.
```

__1d__. What is the difference between  `[` and `[[` subsetting on lists?

```{r q_1d, eval = FALSE}
##`[[` only returns one element, and `[` may return a list of elements. One bracket give a slice of the data, and double bracket goes deeper to retrieve the most granular element.
```

__1e__. What's the difference between `&` and `&&`?

```{r q_1e, eval = FALSE}
##'&&' evaluates and returns a single element (left-to-right associativity), while '&' evaluates and returns multiple elements (ie. vector). 
```

__1f__. Look up the definition of the function `is.primitive`
     and explain _precisely_ how it works.

```{r q_1f, eval = FALSE}
?is.primitive
is.primitive(typeof(4))
##evaluates whether an argument is a primitive function by determining whether the function is found at the base package. It then returns a boolean (T/F).
```

__1g__. How does the `save/load` pair of functions differ from
`saveRDS/readRDS` pair?

```{r q_1g, eval = FALSE}
##save/load saves and loads many objects, while `saveRDS/readRDS`only saves one object at a time.
```

__1h__. Download the source code for R version 3.6.2 from CRAN.  In what file is
    the source for the R function `readBin` located
    _relative_ to the top-level directory `R-3.6.2`?

```{r q_1h, eval = FALSE}
## src > library > base > man
```

__1i__. Why is the comment `##hence length(what) == 1` correct
    in the source for the R function `readBin`?
```{r q_1i, eval = FALSE}
##These are functions to manipulate connections, and the conditions of  "what" are checked such that !is.character(what) || is.na(what) || length(what) != 1L . This means that length must be 1.
```

__1j__. What two features of R make arguments such as `as.is=!stringsAsFactors, fill=!blank.lines.skip` work in
    `read.table` ?

```{r q_1j, eval = FALSE}
##lazy evaluation and argument matching
```

__1k__. Explain the result of
```{r, eval = FALSE}
f1 <- function(x = {y <- 7; 2}, y = 0) {
x + y
print(x)
print(y)
}
f1()


```
in two or three sentences. 

```{r q_1k, eval = FALSE}
##A custom function is created as 'f1': Variables 'x' and 'y' are added. Variable 'y' is called into scope of the variable 'x' equation and is computed as a variable (default value set to 7). X represents the value 2. The function is called in line #86. Result is 9.
```

## Session Info

Leave everything below as is.

```{r}
sessionInfo()
```
