---
title: 'Stat 290: Assignment 2: Q3'
output:
  html_document:
    df_print: paged
---
```{r, echo = FALSE, message=FALSE}
library(tidyverse)
```

## Q3. Tidy data, `dplyr` manipulations and summaries [15 pts]

__3.__ For this exercise, we will use the data from the [2018
  Statistical Computing
  Expo](http://community.amstat.org/stat-computing/data-expo/data-expo-2018).
  The data is provided for you with the assignment in `expo2018.zip`,
  so no need to download it.  In this exercise, we will try to answer
  some of the questions posed in the `Description.txt` file. 

__3a.__ Using `readr` functions, read the three datasets and name them
`locations`, `forecast` and `histWeather` matching the file
names. Names for `forecast` should be changed to pass the check
below. _Hint:_ use `readr::read_csv` and `readr::read_table2` to read
and `readr::problems` on read value to identify issues to fix.

Use the following names for `forecast`:

- `cityRow` for the city row
- `fDate` for the forecasted date
- `fValue` for the forecasted value
- `fWhat` for the forecasted measure
- `fWhen` for the date the forecast was made.


```{r q_3a, eval = TRUE}
##Your ANSWER HERE
locations <- readr::read_csv("locations.csv")
histWeather <- readr::read_csv("histWeather.csv")
col_types <- list(cityRow = readr::col_integer(),
                  fDate = readr::col_date(format = "%Y-%m-%d"),
                  fValue = readr::col_integer(),
                  fWhat = readr::col_character(),
                  fWhen = readr::col_date(format = "%Y-%m-%d"))

forecast <- readr::read_table2( "forecast.dat",
                               col_names = names(col_types),
                               col_types = do.call(readr::cols, col_types),
                               progress = FALSE,
                               na = "M")


```

__3b.__ Ensure `histWeather` is tidy. Mutate `forecast` by removing
missing forecast values.

```{r q_3b, eval = TRUE}
##Your ANSWER HERE
histWeather <- histWeather %>%
    group_by(AirPtCd, Date) %>%
    filter(row_number() == 1)

forecast <- forecast %>%
    filter(!is.na(fValue))

```

__3c.__ Create a dataset to answer the following question: what is
the distribution of errors for maximum temperature in the forecast for
each city? 

For this part, all you do is create a dataset named `maxT` with the
appropriate contents. It should have an integer variable named `lag`
(_date when the forecast was made_ - _forecasted date_) as a measure
of closeness of the forecast, and variable names `trueValue` for the
value from `histWeather` for maximum temperature. Clean out missing
values.

```{r q_3c, eval = TRUE}
##Your ANSWER HERE
## Add column cityRow to locations
locations <- locations %>%
    mutate(cityRow = row_number())

maxT <- forecast %>%
    dplyr::filter(fWhat == "MaxTemp" & !is.na(fValue)) %>%
    dplyr::mutate(lag = as.integer(fWhen - fDate)) %>%
    left_join(locations, by = "cityRow") %>%
    select(fDate, fValue, lag, cityRow, AirPtCd) %>%
    left_join(histWeather, by = c("fDate" = "Date", "AirPtCd" = "AirPtCd")) %>%
    select(cityRow, fDate, fValue, lag, trueValue = "Max_TemperatureF") %>%
    filter(!is.na(trueValue))

```

__3d.__ How do the errors vary with the closeness of the forecast for
	each city? For this part, summarize using mean of
	$\sum_{i=1}^n(y_{f} - y_{h})^2$ as the error measure. (Here, $y_f$
	and $y_h$ are the forecast and true historical value
	respectively). Focus only on forecasts made 2 days before or
	earlier. Your answer should be a grouped dataset named `errorT`
	with columns `cityRow`, `lag` and `error` and group `cityRow`.


```{r q_3d, eval = TRUE}
##Your ANSWER HERE
errorT <- maxT %>%
    filter(lag < -1) %>%
    group_by(cityRow, lag) %>%
    summarize(error = mean((fValue - trueValue)^2))

```

__3e.__ For each city which lag produced the least error? Answer
should be a dataset named `minErrorT` with `city`, `state`, and `lag`
and `error` as variables.

```{r q_3e, eval = TRUE}
##Your ANSWER HERE
minErrorT <- errorT %>%
    slice(which.min(error)) %>%
    left_join(locations, by = "cityRow") %>%
    select(city, state, lag, error)

```
