---
title: 'Stat 290: Assignment 2: Q1'
output:
  html_document:
    df_print: paged
---

## Q1 (Short answers) [10 pts]

__1.__ In class, we showed how to exploit scoping rules (see
  `lec03-md/3.4-State-Preservation.html`) and created the function
  `makePerson`. Here is the excerpted relevant part.

```{r, eval = FALSE}
makePerson <- function(lastName = "", firstName = "") {
  getLastName <- function() lastName
  getFirstName <- function() firstName
  setLastName <- function(what)
      lastName <<- what
  setFirstName <- function(what)
      firstName <<- what
  getFullName <- function()
      paste(lastName, firstName, sep = ", ")
  list(getFullName = getFullName,
       getFirstName = getFirstName,
       getLastName = getLastName,
       setFirstName = setFirstName,
       setLastName = setLastName)
}

```

__1a.__ What are the local variables to `makePerson`? Your answer should
be a character vector of the local variables like `c("a", "b")` with
`a` and `b` replaced appropriately. Empty vectors should be `c()`.
```{r q_1a, eval = TRUE}
##Your ANSWER HERE
c("getFullName",
  "getLastName",
  "getFirstName",
  "setLastName",
  "setFirstName"
  )

```

__1b.__ What are the bound promise variables to `makePerson`? Same
answer format as 1a.
```{r q_1b, eval = TRUE}
##Your ANSWER HERE
c("lastName",
  "firstName"
  )

```

__1c.__ What are the local variables to `setLastName`?
```{r q_1c, eval = TRUE}
##Your ANSWER HERE
c()
```

__1d.__ What are the bound promise variables to `setLastName`?
```{r q_1d, eval = TRUE}
##Your ANSWER HERE
c("what")
```

__1e.__ The above code can be used as follows:

```{r, eval = FALSE}
p1 <- makePerson("Blow", "Joe")
p1$setLastName("Sixpack")
p1$getLastName()
```

Now suppose `makePerson` is changed as below; note the change of args
and calls to `setLastName` and `setFirstName`.

```{r, eval = FALSE}
makePerson <- function(x = "", y = "") {
    getLastName <- function() lastName
    getFirstName <- function() firstName
    setLastName <- function(what)
        lastName <<- what
    setFirstName <- function(what)
        firstName <<- what
    getFullName <- function()
        paste(lastName, firstName, sep = ", ")
    setLastName(x)
    setFirstName(y)
    list(getFullName = getFullName,
         getFirstName = getFirstName,
         getLastName = getLastName,
         setFirstName = setFirstName,
         setLastName = setLastName)
}
```

Consider what happens when we use it in an R session, just as before.

```{r, eval= FALSE}
p2 <- makePerson("Blow", "Joe")
p2$setLastName("Sixpack")
p2$getLastName()
```

Are the above two versions of `makePerson` equivalent? In other words,
do the sequence of calls for `p1` and `p2` have the same effect?
Answer `TRUE` if yes, `FALSE` if no. (Not quoted `"TRUE"` or
`"FALSE"`, just plain R values, without quotes.)


```{r q_1e, eval = TRUE}
##Your ANSWER HERE
FALSE
```

__1f.__ Justify your answer to 1e (1-2 sentences). Your answer should
be text.

```{r q_1f, eval = FALSE}
##Your ANSWER HERE
"You do get the expected answer on the surface 'Sixpack' but a variable
lastName got created in the global environment!"
```

__Note to TAs__ This part of the answer below is only for explanation,
the above answer will suffice from the students!.

In general, the above is dangerous, because you `p2` will not behave
predictably; changes in the global environment---which are
inevitable---will change what happens as the example below shows.

```{r, eval = FALSE}
lastName <- "I can trip you up badly."
p2$getLastName()
```

The moral of the story here is: `<<-` should be used judiciously where
the scope can be carefully controlled! 
