---
title: "20.2. Web API Calls in R (Modified)"
output: slidy_presentation
---

```{r, echo = FALSE, message = FALSE}
## Install a package if not already installed
installIfNeeded <- function(packages, ...) {
    toInstall <- packages[!(packages %in% installed.packages()[, 1])]
    if (length(toInstall) > 0) {
        install.packages(toInstall, repos = "https://cloud.r-project.org")
    }
}

## Ensure packages exist and activate them
needPackages <- function(packages) {
    installIfNeeded(packages)
    for (x in packages) {
        library(x, character.only = TRUE)
    }
}
needPackages(c("tidyverse"))
```


## 20.2. Introduction

Much of the world today relies on data that is made available online
via Application Programming Interfaces (APIs). For example, Github
makes available a set of feeds from which data can be obtained. 

A bit of terminology.

- The entity that accepts requests is called the _server_
- The entity that makes the requests is called the _client_.

___

Examine [Github
Feeds](https://developer.github.com/v3/activity/feeds/). Things to
note:

- There are several feeds
- Some are public and some require authentication
- The results are JSON

Focusing on the `Events` link on the right, you can see that you can
get details on several types of events. 

For example, to obtain basic list of public events, the call is to a
specific form of a URL. 

```
GET /events
```

The part following the verb (`/events`) is called an _end-point_ and
they are the fundamental concepts in RESTful APIs. _REST_ here is an
acronym for (Representational State Transfer) that web applications
use with HTTP protocol. 

For us, it is enough to know that the HTTP protocol consists of
specific _verbs_:

- GET for looking up a resource
- PUT, POST, DELETE for modifying, creating, and deletion requests

Rarely will you use PUT or DELETE, so we focus on GET and POST.

An important detail is that web apis are _stateless_: each
request/response is handled on its own.

___

It is instructive to understand this all in the context of your
browser. For example, in Chrome, you can choose "Developer Tools" and
use the "Network" tab to see what requests are being made by your
browser.  

Let's examine the [Github website](https://github.com).  We will also
use a command line client called `curl` that comes standard with MacOS
to examine the content. (You can download that for Windows, or also
use `wget` for example).

To download the Github public events, we try:

```{bash, eval = FALSE}
curl -o events.json https://api.github.com/events
```

Notice how the above Github endpoint got prefixed with
`https://api.github.com` to refer to an actual site.

___

The previous example was a fixed URL, with no real`a
flexibility. Consider now, the Open FDA website that provides adverse
event reports on drugs. This is described
[here](https://open.fda.gov/apis/try-the-api/).


Suppose we want, say 5, adverse reports for today. 


```{bash, eval = FALSE}
curl -o fda.json 'https://api.fda.gov/drug/event.json?limit=3'
```

Here we are asking for a maximum of 3 reports (and therefore, we can
pass parametrize our requests). 

```{r}
library(jsonlite)
d <- fromJSON('https://api.fda.gov/drug/event.json?search=&limit=3')
```
___

```{r}
str(d)
```

___

Thus, you can programmatically construct a URL to call that API end
point.  Our goal is to do this in R.

In each case, the client `curl` or `R` made the request.

Then the server responded with something and sent data back to the
server. The data it responded with was JSON.

So these are two basic components of the There are two important parts
to http: 

- the __request__, the data sent to the server
- the __response__, the data sent back from the server. 

The request can include parameters. In the case of simple parameters,
these can be passed along with GET requests by means of the `?` and
`&` combination. 

However, if we have a large number of parameters, or if the parameter
to be sent is a data set for example, then the GET request becomes
unwieldy. That is also the case when the parameter may be an image,
which is binary data for example. There is no way to type out the
image as a bunch of text in a URL.

Our goal is to be able to perform all these activities in R.

## 20.2.2 The `httr` Package

What follows is almost entirely from the excellent set of `httr`
vignettes by Hadley Wickham.

To make a request, first load httr, then call `GET()` with a url:

```{r}
library(httr)
r <- GET("http://httpbin.org/get")
```

This gives you a response object. Printing a response object gives you
some useful information: the actual url used (after any redirects),
the http status, the file (content) type, the size, and if it's a text
file, the first few lines of output.

```{r}
r
```

You can pull out important parts of the response with various helper
methods, or dig directly into the object:

```{r}
status_code(r)
headers(r)
glimpse(content(r))
```

## 20.2.3. The Response 

The data sent back from the server consists of three parts: 

- the status line
- the headers and 
- the body. 

The most important part of the status line is the http status code: it
tells you whether or not the request was successful. 


## 20.2.4. The Status Code

The status code is a three digit number that summarises whether or not
the request was successful (as defined by the server that you're
talking to). You can access the status code along with a descriptive
message using `http_status()`:


```{r}
r <- GET("http://httpbin.org/get")
# Get an informative description:
http_status(r)

# Or just access the raw code:
r$status_code
```

A successful request always returns a status of 200. 

Common errors are 400 (bad request), 404 (file not found) and 403
(permission denied).

You can automatically throw a warning or raise an error if a request
did not succeed:

```{r}
warn_for_status(r)
stop_for_status(r)
```

These are very useful when `httr` is used inside a function (i.e. not
interactively) to make sure you find out about errors as soon as
possible.

## 20.2.5. The Body

This is the part that is interesting to the data consumer. Access body
of request using `content()`:

*   `content(r, "text")` accesses the body as a character vector:

    ```{r}
    r <- GET("http://httpbin.org/get")
    content(r, "text")
    ```

    httr will automatically decode content from the server using the encoding 
    supplied in the `content-type` HTTP header. Unfortunately you can't always 
    trust what the server tells you, so you can override encoding if needed:

    ```{r, eval = FALSE}
    content(r, "text", encoding = "ISO-8859-1")
    ```

    If you're having problems figuring out what the correct encoding 
    should be, try `stringi::stri_enc_detect(content(r, "raw"))`.

*   For non-text requests, you can access the body of the request as a 
    raw vector:

    ```{r}
    content(r, "raw")
    ```
    
    This is exactly the sequence of bytes that the web server sent, so this is
    the highest fidelity way of saving files to disk:
    
    ```{r, eval = FALSE}
    bin <- content(r, "raw")
    writeBin(bin, "myfile.txt")
    ```

*   httr provides a number of default parsers for common file types:

    ```{r}
    # JSON automatically parsed into named list
    str(content(r, "parsed"))
    ```
    
    See `?content` for a complete list.
    
    These are convenient for interactive usage, but if you're writing an API
    wrapper, it's best to parse the text or raw content yourself and check it
    is as you expect. See the API wrappers vignette for more details.


## 20.2.6. The headers

Access response headers with `headers()`:

```{r}
headers(r)
```

This is basically a named list, but because http headers are case
insensitive, indexing this object ignores case:

```{r}
headers(r)$date
headers(r)$DATE
```

## 20.2.7. Cookies

Access cookies in a similar way:

```{r}
r <- GET("http://httpbin.org/cookies/set", query = list(a = 1))
cookies(r)
```

Cookies are automatically persisted between requests to the same
domain:

```{r}
r <- GET("http://httpbin.org/cookies/set", query = list(b = 1))
cookies(r)
```

## 20.2.8. The Request

Like the response, the request consists of three pieces: 

- a status line
- headers and 
- a body. 

The status line defines the http method (GET, POST, etc) and
the url. 

You can send additional data to the server in the url (with the query
string), in the headers (including cookies) and in the body of
`POST()` for example.

## 20.2.9. The URL Query String

A common way of sending simple key-value pairs to the server is the
query string: e.g. `http://httpbin.org/get?key=val`. 

`httr` allows you to provide these arguments as a named list with the
`query` argument. 

For example, if you wanted to pass `key1=value1` and `key2=value2` to
`http://httpbin.org/get` you could do:

```{r}
r <- GET("http://httpbin.org/get", 
  query = list(key1 = "value1", key2 = "value2")
)
content(r)$args
```

___

### Example 

For the Open FDA example with 5 records:

```{r}
r <- GET("https://api.fda.gov/drug/event.json/", 
         query = list(search = "",  limit = "5")
         )
```

If we wished to do process this data, we do do the following.

```{r}
ae_content  <- content(r)
records  <- ae_content$results

print(length(records))  ## should be 5
```

Now we just gather the data in a format that works for us.

```{r}

patient_ids  <- 1:5  ## We make up ids
extract_patient_info  <- function(id, records) {
    patient  <- records[[id]]$patient
    patient_age  <- patient$patientonsetage
    patient_sex = patient$patientsex
    ## Patient could have many reactions
    reactions  <- lapply(patient$reaction, function(x) x$reactionmeddrapt)
    product <- patient$drug[[1]]$medicinalproduct
    ## Return a list of things we care about
    list(id = id, age = patient_age, sex = patient_sex,
         drug_reactions = reactions, drug_product = product)
}

## Adverse events info
ae_info  <- lapply(patient_ids, extract_patient_info, records = records)
```

We can take a look at the results

```{r}
str(ae_info)
```

___

Note that `httr::GET` will handle `NULL` elements and quote things
appropriately: any `NULL` elements are automatically dropped from the
list, and both keys and values are escaped automatically.

```{r}
r <- GET("http://httpbin.org/get", 
  query = list(key1 = "value 1", "key 2" = "value2", key2 = NULL))
content(r)$args
```

Note the format of `key 2` which has spaces in it.


## 20.2.10. Request body

When `POST()`ing, you can include data in the `body` of the
request. `httr` allows you to supply this in a number of different
ways. The most common way is a named list:

```{r}
r <- POST("http://httpbin.org/post", body = list(a = 1, b = 2, c = 3))
```

You can use the `encode` argument to determine how this data is sent to the server:

```{r}
url <- "http://httpbin.org/post"
body <- list(a = 1, b = 2, c = 3)

# Form encoded
r <- POST(url, body = body, encode = "form")
# Multipart encoded
r <- POST(url, body = body, encode = "multipart")
# JSON encoded
r <- POST(url, body = body, encode = "json")
```

To see exactly what's being sent to the server, use
`verbose()`. Unfortunately due to the way that `verbose()` works,
knitr can't capture the messages, so you'll need to run these from an
interactive console to see what's going on.

```{r, eval = FALSE}
POST(url, body = body, encode = "multipart", verbose()) # the default
POST(url, body = body, encode = "form", verbose())
POST(url, body = body, encode = "json", verbose())
```

___

You can also send files off disk:

```{r, eval = FALSE}
POST(url, body = upload_file("mypath.txt"))
POST(url, body = list(x = upload_file("mypath.txt")))
```

(`upload_file()` will guess the mime-type from the extension - using
the `type` argument to override/supply yourself.)

These uploads stream the data to the server: the data will be loaded
in R in chunks then sent to the remote server. This means that you can
upload files that are larger than memory.

See `POST()` for more details on the other types of thing that you can
send: no body, empty body, and character and raw vectors.

We will also see further examples of `POST` with R web services.

## 20.2.11. Session Info

```{r}
sessionInfo()
```

